### 2024-02-11 homework_clean.Rmd
added: draft code for exercise 1

---
title: "MAP566 - Homework assignment #1"
format:
  html:
    self-contained: true
    theme: [cosmo, theme.scss]
    toc: true
    number-sections: true
    html-math-method: katex
    code-copy: true
    code-summary: "Show the code"
    code-overflow: wrap
---

# Single comparison

```{r,warning=FALSE}
library(readr)
library(dplyr)
library(tidyverse)

NHANES <- read_csv("NHANES_age_prediction 3.csv")
data = NHANES[,c("DIQ010","age_group","RIDAGEYR","RIAGENDR","PAQ605","BMXBMI","LBXGLU","LBXGLT","LBXIN")]
colnames(data) = c("Diabete","age_group","Age","Sex","Phys_activ","BMI","Glu","Glu2h","BIL")
```

We study data from the National Health and Nutrition Examination Survey (NHANES), administered by the Centers for Disease Control and Prevention (CDC), which collects extensive health and nutritional information from a diverse U.S. population.

```{r}
rmarkdown::paged_table(data)
```

The variables under study are :

-   `Diabete`: diabete diagnosis (1: Yes, 2: No, 3: Borderline);
-   `age_group` : age group of respondent ("Adult" or "Senior");
-   `Age` : the age of respondent;
-   `Sex`: sex of respondent (1: Male, 2: Female);
-   `Physical activity`: respondent's answer to the question "Does your work involve vigorous-intensity activity that causes large increases in breathing or heart rate like carrying or lifting heavy loads, digging or construction work for at least 10 minutes continuously? (1: Yes, 2: No, 7: no answer);
-   `BMI` : Body Mass Index if respondent ($kg/m^{2}$);
-   `Glu` : Respondent's Blood Glucose after fasting ($mg/dL$);
-   `Glu2h` : Respondent's Two Hours Blood Glucose ($mg/dL$);
-   `BIL` : Respondent's Blood Insulin Level ($pmol/L$).

1.  Test if the mean level of blood glucose after fasting is the same for diabetic and non diabetic respondents (*hint:* plot first the data and justify the test(s) to use).

    To investigate if the mean level of blood glucose after fasting is the same for diabetic and non-diabetic we should first split the data and plot the distributions side by side to get an idea for which test will be the most appropriate.

    ```{r}
    diabetic <- data %>% filter(Diabete == 1)
    non_diabetic <- data %>% filter(Diabete == 2)

    rmarkdown::paged_table(diabetic)
    rmarkdown::paged_table(non_diabetic)
    ```

    ```{r}
    diabetic$Group <- "Diabetic"
    non_diabetic$Group <- "Non-Diabetic"

    combined_data <- rbind(diabetic, non_diabetic)

    ggplot(combined_data, aes(x = Group, y = Glu)) +
      geom_boxplot() +
      labs(x = "Status", y = "Glucose") +
      ggtitle("Glucose Distribution by Diabetes Status") +
      theme_minimal()

    ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
      geom_violin(trim = FALSE) +
      labs(x = "Status", y = "Glucose") +
      ggtitle("Glucose Distribution by Diabetes Status") +
      theme_minimal()
    ```

    It is clear that the glucose distribution after fasting for the two groups have vastly different variances, thus a two-sample t-test would not be applicable due to its equal variance assumption. Further, it appears that the two distributions follow somewhat the normal assumption. **However, we do observe that the diabetic distribution has an elongated tail towards higher glucose levels which could be breaking this assumption.**

    [**TODO :: Double check that indeed the diabetic distribution follows the normal assumption**]{.underline}

    ```{r}
    var.test(diabetic$Glu, non_diabetic$Glu, var.equal=TRUE)
    ```

    Further, it is now clear that indeed the variances are not the same for the two distributions given that we can clearly reject the null hypothesis stating that the two variances are the same.

    We will thus be using a Welch two sample t-test to examine whether or not the means of the two groups are the same given that this test does not assume equal variance.

    [**TODO :: Add the setup for the null hypothesis**]{.underline}

    ```{r}
    t.test(diabetic$Glu, non_diabetic$Glu)
    ```

    [**TODO :: Write conclusion about the results observed**]{.underline}

2.  Test for the diabetic respondents if the mean level of blood glucose after fasting is the same for adults and seniors.

    Again we need to investigate the distribution for the two groups to understand which assumptions for the different hypothesis tests will be applicable.

    ```{r}
    diabetic_adult <- data %>% filter(Diabete == 1 & age_group == "Adult")
    diabetic_seniors <- data %>% filter(Diabete == 1 & age_group == "Senior")

    rmarkdown::paged_table(diabetic_adult)
    rmarkdown::paged_table(diabetic_seniors)
    ```

    ```{r}
    diabetic_adult$Group <- "Diabetic Adult"
    diabetic_seniors$Group <- "Diabetic Senior"

    combined_data <- rbind(diabetic_adult, diabetic_seniors)

    ggplot(combined_data, aes(x = Group, y = Glu)) +
      geom_boxplot() +
      labs(x = "Age Group", y = "Glucose") +
      ggtitle("Glucose Distribution for Diabetics by Age Group") +
      theme_minimal()

    ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
      geom_violin(trim = FALSE) +
      labs(x = "Age Group", y = "Glucose") +
      ggtitle("Glucose Distribution for Diabetics by Age Group") +
      theme_minimal()
    ```

    From this we can observe that the distribution for the diabetic adults and diabetic seniors do not have the same variance and that they do not follow the normal assumption. Which is clear by the fact that for seniors we have two distinct peaks, and for adults we have a trailing tail for higher glucose levels which indicate that they do not follow the parametric Gaussian distribution. Making the two sample Welch t-test not applicable in this scenario. Thus, to circumvent these limiting assumptions in this case we will be performing a Wilcoxon test which does not have an underlying parametric assumption about the distribution of the data.

    [**TODO :: Add the maths behind to frame the problem first**]{.underline}

    ```{r}
    wilcox.test(diabetic_adult$Glu, diabetic_seniors$Glu, alternative="two.sided")
    ```

    [**TODO :: Add conclusion about the results of the test**]{.underline}

3.  Is it possible to test for the adults diabetic respondents if the mean level of blood glucose after fasting is the same for those who have a vigorous work activity and for those who have not?

    [**TODO :: Write motivation for the test**]{.underline}

    ```{r}
    diabetic_active <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 1)
    diabetic_inactive <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 2)

    diabetic_active$Group <- "Diabetic Active"
    diabetic_inactive$Group <- "Diabetic Inactive"

    # Combine the data frames
    combined_data <- rbind(diabetic_active, diabetic_inactive)

    # Create boxplot
    ggplot(combined_data, aes(x = Group, y = Glu)) +
      geom_boxplot() +
      labs(x = "Work Activity", y = "Glucose") +
      ggtitle("Glucose Distribution for Diabetic Adults by Work Activity") +
      theme_minimal()

    # Create violin plot
    ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
      geom_violin(trim = FALSE) +
      labs(x = "Work Activity", y = "Glucose") +
      ggtitle("Glucose Distribution for Diabetic Adults by Work Activity") +
      theme_minimal()
    ```

    ```{r}
    wilcox.test(diabetic_active$Glu, diabetic_inactive$Glu, alternative="two.sided")
    ```

    [**TODO :: Add conclusion and analysis about the results**]{.underline}

4.  Test if the proportion of diabetic is the same for male and female respondents. Compare conclusions of several tests. You can use the function `table` to compute contingency tables.

    ```{r}
    # Create a contingency table
    contingency_table <- table(data$Diabete, data$Sex)
    rownames(contingency_table) <- c("No Diabetes", "Diabetes", "Borderline")
    colnames(contingency_table) <- c("Male", "Female")

    # Display the contingency table
    contingency_table

    # Perform chi-square test
    chi_square_test <- chisq.test(contingency_table)

    # Display the results
    chi_square_test
    ```

# Gene expression data

The *liver* dataset contains measurements of rat liver toxicity levels (measured through cholesterol levels) as well as measurements of the expression levels of several thousand genes.

When loading the data (file `liver_data.rda`), the table *liver* is created with 64 rows (the observations) and 3117 columns. The data can be loaded using the following code:

```{r}
load("liver_data.rda")
```

The first column, `cholesterol`, is the variable to be explained. The remaining 3116 are the expressions of 3116 genes (more precisely, the logarithm of the ratio between expression levels in two experimental conditions). The aim is to identify the variables (and therefore the genes) linked to the response.

1.  We first want to study the gene which is the most correlated with the level of cholesterol. Identify this gene and plot `cholesterol` as a function of this gene expression level. Fit a linear model. Test if there is a significant relation between this gene expression and `cholesterol`?

2.  We now wish to perform the previous test for all of the 3116 genes. For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the $p$-value of the test.

3.  Order the $p$-values and plot the ordered $p$-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.

4.  Identify a set of genes linked to the response (*aka* discoveries). We want to guarantee that the expected proportion of false discoveries (mistakes) is less than $5\%$. Explain how you proceed and how many genes you discover.

5.  We wish to be more conservative and guarantee that the probability of making a false discovery (or more) is less than $5\%$. Explain how you proceed and how many genes you discover.

# Non parametric regression

1.  Upload the dataset contained in the file `data_exo3.csv` and plot the data.

2.  Try several polynomial models for the data, select a model and comment the results.

3.  Fit a nonlinear model and compare with the polynomial model you have selected.

4.  Calculate confidence intervals for $\mathbb{E}(Y_{new})$ and prediction intervals for $Y_{new}$ on the following values of the covariate: $$
    x_{new} = (1,...,10). 
    $$

# S&P500 daily return

The Standard and Poor's 500 or simply the S&P 500, is a stock market index tracking the performance of 500 large companies listed on stock exchanges in the United States. It is one of the most commonly followed equity indices.

The dataset `sp500_history.csv` contains information about the value of the index at the opening and closing of the market, every day from January 3rd, 2007 to February 2nd, 2024. The goal of the study is to understand the distribution of the *Daily Return* during this period. The daily return is defined as the index difference between closing and opening.

1.  Load the dataset `sp500_history.csv` into a *data frame* and add a column that computes the daily return for each days.

2.  We propose to model the daily return as a sample from a normal population. Write the model and use *R* to fit it to the data. What do you think of this model?

3.  Instead of a single normal, we now propose to use a mixture of normals with $p$ components. Write the corresponding models and use *R* to fit them to the data for all values of $p$ between $2$ and $6$. Discuss the results and compare them with the model of Question 1.

4.  As a third and last model, we propose to model the daily return as coming from a location-scale family of Student distributions, which is a model defined by the collection of densities $$f_{\nu,m,a}(x) = \frac{\Gamma\big(\frac{\nu+1}{2}\big)}{\sqrt{\pi \nu a^2}\Gamma\big(\frac{\nu}{2}\big)}\Bigg(1+\frac{(x-m)^2}{\nu a^2} \Bigg)^{-\frac{\nu+1}{2}}$$ where $\Gamma$ is the Gamma function and $\theta=(\nu,m,a) \in \mathbb{R}_+^*\times \mathbb{R}\times \mathbb{R}_+^*$ is the parameter.

    1.  What is the motivation for considering this model?
    2.  Propose an algorithm for estimating the parameter of this model and use it to fit this model to the data.

5.  Between all the previous models (the normal, the 5 mixtures of normals, and the location-scale Student) which one do you choose? Explain your methodology.

6.  Is the expected daily return different than zero?

### 2024-02-13 homework_clean.Rmd
added: draft work on MM section

@@
 library(tidyverse)
+library(MASS)
 
@@
 
+    ```{r}
+    sp500_data <- read_csv("sp500_history.csv")
+    rmarkdown::paged_table(sp500_data)
+    ```
+
+    ```{r}
+    sp500_data$DailyReturn <- sp500_data$Close - sp500_data$Open
+    rmarkdown::paged_table(sp500_data)
+    ```
+
 2.  We propose to model the daily return as a sample from a normal population. Write the model and use *R* to fit it to the data. What do you think of this model?
+
+    ```{r}
+    ggplot(sp500_data, aes(x = Date, y = DailyReturn)) +
+      geom_histogram(stat = "identity", fill = "skyblue", color = "black", bins = 15) +
+      labs(title = "Daily Return of S&P 500 (2007-2010)",
+           x = "Date",
+           y = "Daily Return") +
+      theme_minimal() +
+      theme(axis.text.x = element_blank(),
+            axis.title.x = element_blank())
+    ```
+
+    ```{r}
+    fit <- fitdistr(sp500_data$DailyReturn, "normal")
+
+    # Print the fitted parameters
+    print(fit)
+
+    # Plot the histogram of daily returns with the fitted Gaussian distribution
+    hist(sp500_data$DailyReturn, breaks = 30, freq = FALSE, 
+         main = "Histogram of Daily Returns with Fitted Gaussian Distribution",
+         xlab = "Daily Return")
+    curve(dnorm(x, mean = fit$estimate[1], sd = fit$estimate[2]), 
+          col = "blue", lwd = 2, add = TRUE, yaxt = "n")
+    ```
 

### 2024-03-01 homework_clean.Rmd
feat: method justifications for ex4

--- 2024-02-11 homework_clean.Rmd
@@
 
+    ```{r}
+    sp500_data <- read_csv("sp500_history.csv")
+    rmarkdown::paged_table(sp500_data)
+    ```
+
+    ```{r}
+    sp500_data$DailyReturn <- sp500_data$Close - sp500_data$Open
+    rmarkdown::paged_table(sp500_data)
+    ```
+
 2.  We propose to model the daily return as a sample from a normal population. Write the model and use *R* to fit it to the data. What do you think of this model?
+
+    ```{r}
+    ggplot(sp500_data, aes(x = Date, y = DailyReturn)) +
+      geom_histogram(stat = "identity", fill = "grey", color = "grey", bins = 15) +
+      labs(title = "Daily Return of S&P 500 (2007-2010)",
+           x = "Date",
+           y = "Daily Return") +
+      theme_minimal() +
+      theme(axis.text.x = element_blank(),
+            axis.title.x = element_blank())
+    ```
+
+    Under Louis Bachelier's "Theory of Speculation" assume that the returns can be sampled from a perfectly random distribution, in this case this would be a normal distribution which we can parametirise through a Gaussian model.
+Assume that the return is modeled by some random variable $X_r$ and that some value of that RV is denoted by $x_t$.
+Thus, since we first assume Bachelier's theory we can model the RV through the following Gaussian.
+
+    $$
+    X_r \sim \mathcal{N}(\mu_r, \sigma_r^2)
+    $$
+
+    Where the model parameters are the empirical mean and empirical variance expressed by:
+
+    $$
+    \mu_r = \dfrac{\sum_{i=0}^N x_i}{N} \hspace{5mm} \sigma_r^2 = \dfrac{\sum_{i=0}^N (x_i - \mu_r)}{N}
+    $$
+
+    We can then directly fit such a Gaussian model in R through the following:
+
+    ```{r}
+    fit <- fitdistr(sp500_data$DailyReturn, "normal")
+
+    # Print the fitted parameters
+    print(fit)
+
+    # Plot the histogram of daily returns with the fitted Gaussian distribution
+    hist(sp500_data$DailyReturn, breaks = 100, freq = FALSE, 
+         main = "Histogram of Daily Returns with Fitted Gaussian Distribution",
+         xlab = "Daily Return")
+    curve(dnorm(x, mean = fit$estimate[1], sd = fit$estimate[2]), 
+          col = "blue", lwd = 2, add = TRUE, yaxt = "n")
+    ```
+
+    Indeed, the it seems that Bachelier's theory does describe the price movements well. As the distribution of the returns follow quite clearly a normal distribution.
 
@@
 Discuss the results and compare them with the model of Question 1.
+
+    To combine the multiple Gaussians to model a single valid PDF, we will have to introduce a weight parameter which will be used between the weighted sum of Gaussian components.
+First, we denote a single component the following way.
+
+    $$
+    Y_r \sim \mathcal{N}(\theta_k) \hspace{5mm} \text{Where $\theta$ is a parameter vector for component $k$ defined by} \hspace{5mm} \theta_k = \left[ \mu_k, \sigma_k^2 \right]
+    $$
+
+    Furthermore, to form the mixture we introduce the weight parameter $\pi_k$ which is probabilistic and forms the weighted sum (GMM).
+
+    $$
+    Y_i = \sum_{k=1}^K \pi_k \mathcal{N}(\theta_k)
+    $$
+
+    As this is a weighted sum and we expect $Y_i$ to be modeled from a valid PDF, we have to constrain the weights such that:
+
+    $$
+    \pi_k \geq 0 \hspace{2.5mm} \land \hspace{2.5mm} \sum_{k=1}^K \pi_k = 1
+    $$
+
+    Which is the case as long as we follow that $\pi_k$ comes from a PDF defined over the categorical RVs $Z_i$ which defines the different components such that we have:
+
+    $$
+    P(Z_i = k) = \pi_k \hspace{5mm} \text{Where $Z$ is a sequence of RVs} \hspace{5mm} \{1, 2, \dots, K\}
+    $$
+
+    Thus, we observe that we first have to initialize the parameters of the model which we can do by placing the means based on a K-Means initialization and set some explicit variance which will serve as the initialization for the iterative EM-algorithm for parameter optimization over such a GMM.
 

### 2024-03-02 homework_clean.Rmd
feat: q4 code finished + constrained optimisation

@@
 library(MASS)
+library(mixtools)
 
@@
 sp500_data <- read_csv("sp500_history.csv")
-    rmarkdown::paged_table(sp500_data)
-    ```
+sp500_data$DailyReturn <- sp500_data$Close - sp500_data$Open
 
-    ```{r}
-    sp500_data$DailyReturn <- sp500_data$Close - sp500_data$Open
+# Check for non-numeric values in DailyReturn column
+non_numeric_count <- sum(is.na(sp500_data$DailyReturn) | !is.numeric(sp500_data$DailyReturn))
+
+# Print the count of non-numeric values
+cat("Number of non-numeric values in DailyReturn column:", non_numeric_count, "\n")
+
+# If there are non-numeric values, print their indices
+if (non_numeric_count > 0) {
+  cat("Indices of non-numeric values in DailyReturn column:", which(!is.numeric(sp500_data$DailyReturn)), "\n")
+} else {
+  cat("No non-numeric values found in DailyReturn column.\n")
+}
+
 rmarkdown::paged_table(sp500_data)
@@
 
-    Indeed, the it seems that Bachelier's theory does describe the price movements well. As the distribution of the returns follow quite clearly a normal distribution.
+    Indeed, the it seems that Bachelier's theory does describe the price movements well.
+As the daily return resembles a normal distribution, however, we do observe that that tails seem to follow a higher variance compared to the density around the mean which seems to have a smaller variance.
+Which also explains why our initial Gaussian doesn't effectively capture this density.
+
+    **TODO :: We should maybe fix the means and use two components as we see the bottom part has a larger variance compared to the middle parts.**
+
+    **TODO :: Compare the convergence graphs for the naive approach with no optimisation constraints with the one where we use a optimisation constrain on the means**
 
@@
 
+    First off, we will naively apply a GMM routine to this problem using expectation maximization which we will compare against an constrained optimization approach where we ensure equal means for all distributions as it seems to only be the variance that's needed to change based on our earlier graph.
+
+    ## Unconstrained GMM Optimization
+
+    ```{r}
+    # First we apply K-Means clustering on the graph to optain initial means
+    n_components = 3
+    y = sp500_data$DailyReturn
+
+    kmeans_out <- kmeans(y, centers = n_components)
+    mu = as.vector(kmeans_out$centers)
+    mu
+    ```
+
+    This results seems strange given that we should want Gaussian components around the same mean.
+However, for this experiment we will naively apply k-Means initialization without any constraints in the optimization and observe what results we obtain.
+
+    ```{r}
+    fit_mixture_model <- function(n_components, y, max_iter = 300) {
+      kmeans_out <- kmeans(y, centers = n_components)
+      mu <- as.vector(kmeans_out$centers)
+      mixture_EM <- normalmixEM(y, mu = mu, k = n_components, arbvar = FALSE, maxit = max_iter)
+
+      log_likelihood <- mixture_EM$loglik
+      print(log_likelihood)
+      result <- list(pi = mixture_EM$lambda, mu = mixture_EM$mu, sigma = mixture_EM$sigma)
+
+      result <- list(pi = mixture_EM$lambda, mu = mixture_EM$mu, sigma = mixture_EM$sigma)
+
+      # Plot histogram with 50 bars
+      hist(y, breaks = 50, probability = TRUE, col = alpha("skyblue", 0.25), border = alpha("black", 0.25), main = paste("Mixture Model with", n_components, "Components"), xlab = "Daily Return", ylab = "Density")
+
+      # Add density line
+      lines(density(mixture_EM$x), col="red", lwd=2)
+
+      # Plot densities of Gaussian components
+      x <- seq(min(y), max(y), length.out = 1000)
+      for (i in 1:n_components) {
+        curve(result$pi[i] * dnorm(x, mean = result$mu[i], sd = result$sigma[i]), add = TRUE, col = i + 1, lwd=2, lty=2)
+      }
+
+      return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood))
+    }
+
+    ```
+
+    ```{r}
+    mx2 <- fit_mixture_model(2, y)
+    mx3 <- fit_mixture_model(3, y)
+    mx4 <- fit_mixture_model(4, y)
+    mx5 <- fit_mixture_model(5, y)
+    mx6 <- fit_mixture_model(6, y)
+
+    mixtures = c(mx2, mx3, mx4, mx5, mx6)
+    ```
+We can observe that the final fits are indeed much better than the single Gaussian example.
+However, we do observe that there is some overfitting happening at the tails of the distribution for all of the experiments we have run.
+This is caused by the fact that the means are not the same for all the Gaussian components and thus for our next section we will be looking at constrained optimization of this GMM where the means should be the same.
+
+## Constrained GMM Optimisation
+
+```{r}
+dcomponents <- function(theta, x) {
+    mapply(
+      function(pi, mu, sigma) pi * dnorm(x, mu, sigma),
+      theta$pi, theta$mu, theta$sigma,
+      SIMPLIFY = TRUE
+    )
+}
+```
+
+```{r}
+
+M_step_general <- function(tau, x) {
+  pi    <- colMeans(tau)
+  mu    <- colSums(tau * x) / colSums(tau)
+  sigma <- sqrt(colSums(tau * x^2) / colSums(tau) - mu^2)
+  list(pi = pi, mu = mu, sigma = sigma)
+}
+
+M_step_same_means <- function(tau, x) {
+  n <- length(x)
+  K <- ncol(tau)
+  pi    <- colSums(tau) / n
+  muk   <- colSums(tau * x) / colSums(tau)
+  sigma <- sqrt(colSums(tau * x^2) / colSums(tau) - muk^2)
+  wk    <- n * pi / sigma^2
+  mu    <- rep(sum(wk * muk) / sum(wk), K)
+  list(pi = pi, mu = mu, sigma = sigma)
+}
+
+```
+
+```{r}
+
+mixture_gaussian1D <- function(x, theta0, M_step = M_step_general, max_iter = 100, threshold = 1e-6) {
+
+  ## initialization
+  n <- length(x)
+  likelihoods  <- dcomponents(theta0, x) 
+  deviance     <- numeric(max_iter)
+  deviance[1]  <- -2 * sum(log(rowSums(likelihoods)))
+
+  for (t in 1:max_iter) {
+
+    # E step
+    tau <- likelihoods / rowSums(likelihoods)
+    # M step
+    theta <- M_step(tau, x)
+
+    ## Assessing convergence
+    likelihoods   <- dcomponents(theta, x)
+    deviance[t+1] <- - 2 * sum(log(rowSums(likelihoods)))
+
+    ## prepare next iterations
+
+    if (abs(deviance[t + 1] - deviance[t]) < threshold)
+      break
+
+  }
+
+  list(theta = theta, deviance = deviance[t + 1])
+}
+
+```
+
+```{r}
+mu_data = mean(y)
+pi = c(.5, .5)
+means = c(mu_data, mu_data)
+variances = c(10,2)
+
+theta0  <- list(pi=pi, mu=means, sigma=variances)
+out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)
+out_general$theta
+```
+```{r}
+softmax <- function(x) {
+  exp_x <- exp(x)
+  exp_x / sum(exp_x)
+}
+
+constrained_gmm_experiment <- function(y, variances) {
+  n_components = length(variances)
+
+  mu_data = mean(y)
+
+  pi = softmax(rep(1, n_components))
+  means = rep(mu_data, n_components)
+
+  theta0  <- list(pi=pi, mu=means, sigma=variances)
+  out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)
+  out_general$theta
+
+  # Define the data
+  x <- seq(-150, 150, length.out = 1000)
+
+  # Compute the densities for the two components
+  densities <- dcomponents(out_general$theta, x)
+
+  # Calculate the final density of the mixture
+  final_density <- rowSums(sapply(1:length(out_general$theta$pi), function(i) out_general$theta$pi[i] * dnorm(x, mean = out_general$theta$mu[i], sd = out_general$theta$sigma[i])))
+
+  # Plot the data points
+  hist(y, breaks = 50, probability = TRUE, col = alpha("skyblue", 0.25), border = alpha("black", 0.25), main = paste("Mixture Model with", n_components, "Components"), xlab = "Daily Return", ylab = "Density")
+
+  # Plot the densities for each component
+  for (i in 1:n_components) {
+    lines(x, densities[,i], col = rainbow(n_components)[i], lwd=2, lty=2)
+  }
+
+  # Plot the final density of the mixture
+  lines(x, final_density, lwd = 2)
+
+  # Add legend
+  legend_labels <- c(paste0("Component ", 1:n_components), "Mixture")
+  legend("topright", legend = legend_labels, col = c(rainbow(n_components), "black"), lty = 1, lwd = 2, pch = 20)
+}
+```
+
+```{r}
+constrained_gmm_experiment(y, variances = c(10, 5, 7))
+```
+
+    **TODO :: Compare each using BIC**
+
 4.
@@
     1.  What is the motivation for considering this model?
+
+        Indeed from looking at the above examples the final fits are still not great even when we add additional components to account for the variable variance of the distribution.
+But as we observe the distribution it is more clear that it can follow more closely a student t-distribution given its long tails and elongated middle density.
+
     2.  Propose an algorithm for estimating the parameter of this model and use it to fit this model to the data.
+
+```{r}
+# Define log-likelihood function for Student's t-distribution
+log_likelihood_student_t <- function(theta, x) {
+  nu <- theta[1]
+  mu <- theta[2]
+  sigma <- theta[3]
+
+  # Check if sigma is positive
+  if(sigma <= 0) {
+    return(Inf)
+  }
+
+  # Check if nu is at least 1
+  if(nu < 1) {
+    nu <- 1
+  }
+
+  loglik <- sum(dt((x - mu) / sigma, df = nu, log = TRUE)) - length(x) * log(sigma)
+
+  return(-loglik)  # Negative log-likelihood for minimization
+}
+
+# Fit Student's t-distribution model using MLE
+fit_student_t <- function(y) {
+  # Check for non-finite values in the data
+  if (any(!is.finite(y))) {
+    stop("Input data contains non-finite values.")
+  }
+
+  # Remove missing values
+  y <- y[is.finite(y)]
+
+  # Check if there are enough data points
+  if (length(y) < 3) {
+    stop("Insufficient data points for estimation.")
+  }
+
+  # Initial parameter guess
+  init_params <- c(5, mean(y), sd(y))
+
+  # Optimize log-likelihood function
+  opt_result <- optim(init_params, log_likelihood_student_t, x = y, method = "L-BFGS-B",
+                      lower = c(0, -Inf, 0), upper = c(Inf, Inf, Inf))
+
+  # Extract optimized parameters
+  nu <- opt_result$par[1]
+  mu <- opt_result$par[2]
+  sigma <- opt_result$par[3]
+
+  # Return parameter estimates
+  return(list(nu = nu, mu = mu, sigma = sigma))
+}
+
+# Fit Student's t-distribution model to the data
+student_t_params <- fit_student_t(y)
+student_t_params
+```
+
+```{r}
+# Generate data for the PDF using the fitted parameters
+x_range <- seq(min(y), max(y), length.out = 1000)
+pdf_values <- dt((x_range - student_t_params$mu) / student_t_params$sigma, 
+                 df = student_t_params$nu) / student_t_params$sigma
+
+# Create a data frame for plotting
+plot_data <- data.frame(x = x_range, pdf = pdf_values)
+
+# Plot histogram of the data and fitted PDF
+ggplot() +
+  geom_histogram(data = data.frame(y), aes(x = y, y = ..density..), bins = 30, fill = "skyblue", color = "black") +
+  geom_line(data = plot_data, aes(x = x, y = pdf), color = "red", size = 1) +
+  labs(title = "Fitted Student's t-distribution Model",
+       x = "Daily Return",
+       y = "Density") +
+  theme_minimal()
+```
 

### 2024-03-02 submission.Rmd
feat: q4 code finished + constrained optimisation

---
title: "submission"
output: html_document
date: "2024-03-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Single comparison

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2024-03-03 homework_clean.Rmd
feat: move answers to final submission file (note some subparts are not fully done)

@@
     [**TODO :: Write motivation for the test**]{.underline}
+    TODO :: Apply power of t-test such that we can examine if we can test this
+```{r}
+diabetic_active <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 1)
+diabetic_inactive <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 2)
+
+diabetic_active$Group <- "Diabetic Active"
+diabetic_inactive$Group <- "Diabetic Inactive"
+
+# Combine the data frames
+combined_data <- rbind(diabetic_active, diabetic_inactive)
+combined_data
+```
 
@@
 4.  Test if the proportion of diabetic is the same for male and female respondents. Compare conclusions of several tests. You can use the function `table` to compute contingency tables.
+
+TODO :: Compare several tests
 
@@
     ```
+
     We can observe that the final fits are indeed much better than the single Gaussian example.
@@
 ```
+
 ```{r}
@@
 ```{r}
-constrained_gmm_experiment(y, variances = c(10, 5, 7))
+constrained_gmm_experiment(y, variances = c(10, 5))
 ```
 
+```         
 **TODO :: Compare each using BIC**
+```
 

### 2024-03-03 submission.Rmd
feat: move answers to final submission file (note some subparts are not fully done)

@@
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r,warning=FALSE}
+library(readr)
+library(dplyr)
+library(tidyverse)
+library(MASS)
+library(mixtools)
 ```
 
-# Single comparison
+# Statistics in Action - Homework
 
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+**Group:** Max Rehman Linder, Patrick Tourniaire & Antonio Nisi
 
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
+## 1. Single comparison
 
-```{r cars}
-summary(cars)
+```{r}
+# Import dataset
+
+NHANES <- read_csv("NHANES_age_prediction 3.csv")
+data = NHANES[,c("DIQ010","age_group","RIDAGEYR","RIAGENDR","PAQ605","BMXBMI","LBXGLU","LBXGLT","LBXIN")]
+colnames(data) = c("Diabete","age_group","Age","Sex","Phys_activ","BMI","Glu","Glu2h","BIL")
+
+rmarkdown::paged_table(data)
 ```
 
-## Including Plots
+Feature descriptions for the above dataset:
 
-You can also embed plots, for example:
+-   `Diabete`: diabete diagnosis (1: Yes, 2: No, 3: Borderline)
+-   `age_group` : age group of respondent ("Adult" or "Senior")
+-   `Age` : the age of respondent
+-   `Sex`: sex of respondent (1: Male, 2: Female)
+-   `Physical activity`: respondent's answer to the question "Does your work involve vigorous-intensity activity that causes large increases in breathing or heart rate like carrying or lifting heavy loads, digging or construction work for at least 10 minutes continuously?
+(1: Yes, 2: No, 7: no answer)
+-   `BMI` : Body Mass Index if respondent ($kg/m^{2}$)
+-   `Glu` : Respondent's Blood Glucose after fasting ($mg/dL$)
+-   `Glu2h` : Respondent's Two Hours Blood Glucose ($mg/dL$)
+-   `BIL` : Respondent's Blood Insulin Level ($pmol/L$).
 
-```{r pressure, echo=FALSE}
-plot(pressure)
+### 1.1 Mean level of blood glucose after fasting
+
+**Task description:** Test if the mean level of blood glucose after fasting is the same for diabetic and non diabetic respondents (*hint:* plot first the data and justify the test(s) to use).
+
+Before starting to compare the two groups of diabetics and non-diabetics we first split the dataset by filtering for the two cases.
+
+```{r}
+diabetic <- data %>% filter(Diabete == 1)
+non_diabetic <- data %>% filter(Diabete == 2)
+
+rmarkdown::paged_table(diabetic)
+rmarkdown::paged_table(non_diabetic)
+
+# Create a new combined table with only these two groups
+diabetic$Group <- "Diabetic"
+non_diabetic$Group <- "Non-Diabetic"
+
+combined_data <- rbind(diabetic, non_diabetic)
+combined_data
 ```
 
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
+To be able to accurately test the mean blood glucose level for the two groups we have to plot the distribution of this data.
+This will allow us to identify the most applicable test as each hypothesis test have underlying assumptions which the data should be roughly following to achieve an accurate result through our tests.
+
+To achieve this we create a box plot and a violin plot to understand the mean and variance of the data and to see how the data is distributed differently for the two groups through a violin plot.
+
+```{r}
+ggplot(combined_data, aes(x = Group, y = Glu)) +
+  geom_boxplot() +
+  labs(x = "Status", y = "Glucose") +
+  ggtitle("Glucose Distribution by Diabetes Status") +
+  theme_minimal()
+
+ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
+  geom_violin(trim = FALSE) +
+  labs(x = "Status", y = "Glucose") +
+  ggtitle("Glucose Distribution by Diabetes Status") +
+  theme_minimal()
+```
+
+It is clear that the glucose distribution after fasting for the two groups have vastly different variances, thus a two-sample t-test would not be applicable due to its equal variance assumption.
+Further, it appears that the two distributions follow somewhat the parametric assumption.
+As a sanity check we can setup a test of the to distributions to verify that the variances should not be equal.
+
+To achieve this we can perform an F-test which will test whether the variances of the two groups are equal.
+Thus, we will have the following hypotheses (where $d$ represents the diabetics and $\bar{d}$ represents the non-diabetics):
+
+$$
+\begin{align}
+  H_0 &: \sigma_{d} = \sigma_{\bar{d}} \\
+  H_1 &:\sigma_{d} \neq \sigma_{\bar{d}} \hspace{5mm} \text{for a two-tailed F-test}
+\end{align}
+$$
+
+Since, we are only interested to see if the variances are unequal or not we use the two-tailed alternative hypothesis for the F-test.
+
+```{r}
+var.test(diabetic$Glu, non_diabetic$Glu, alternative = "two.sided")
+```
+
+Indeed, we can report a p-value of 2.2e-16 thus, we can reject the null hypothesis and conclude that the variances are indeed different.
+Which leads us to using the Welch student t-test which assumes parametric distribution with two groups of different variances.
+Again we use the same two sided alternative hypothesis but now for the means.
+
+$$
+\begin{align}
+H_0 &: \mu_{d} = \mu_{\bar{d}} \\
+H_1 &: \mu_{d} \neq \mu_{\bar{d}}
+\end{align}
+$$
+
+With the following piece of code we can perform this Welch t-test on the means.
+
+```{r}
+t.test(diabetic$Glu, non_diabetic$Glu, alternative="two.sided")
+```
+
+Given that $\alpha = 0.05$ we can reject the null-hypothesis and say that our results of this test is significant and shows that we do not have enough evidence to conclude that the means between the two groups are indeed the same.
+
+### 1.2 Diabetic mean blood glucose level between adults and seniors
+
+**Task description:** Test for the diabetic respondents if the mean level of blood glucose after fasting is the same for adults and seniors.
+
+To analyse what test will be the most applicable we will first plot the distributions, but to do that we will first need to create these two groups from our dataset.
+
+```{r}
+diabetic_adult <- data %>% filter(Diabete == 1 & age_group == "Adult")
+diabetic_seniors <- data %>% filter(Diabete == 1 & age_group == "Senior")
+
+rmarkdown::paged_table(diabetic_adult)
+rmarkdown::paged_table(diabetic_seniors)
+
+diabetic_adult$Group <- "Diabetic Adult"
+diabetic_seniors$Group <- "Diabetic Senior"
+
+combined_data <- rbind(diabetic_adult, diabetic_seniors)
+rmarkdown::paged_table(combined_data)
+```
+
+We then use the new combined dataframe consisting only of these two groups to plot the distributions.
+
+```{r}
+ggplot(combined_data, aes(x = Group, y = Glu)) +
+  geom_boxplot() +
+  labs(x = "Age Group", y = "Glucose") +
+  ggtitle("Glucose Distribution for Diabetics by Age Group") +
+  theme_minimal()
+
+ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
+  geom_violin(trim = FALSE) +
+  labs(x = "Age Group", y = "Glucose") +
+  ggtitle("Glucose Distribution for Diabetics by Age Group") +
+  theme_minimal()
+```
+
+From this we can observe that the distribution for the diabetic adults and diabetic seniors do not have the same variance and that they do not follow the parametric assumption.
+Which is clear by the fact that for seniors we have two distinct peaks which indicate that they do not follow the parametric distribution.
+Making the two sample Welch t-test not applicable in this scenario.
+Thus, to circumvent these limiting assumptions in this case we will be performing a Wilcoxon test which does not have an underlying parametric assumption about the distribution of the data.
+
+Given the task description we can setup the following hypothesis test. Where $a$ represents the adult group and $s$ represents the senior group.
+
+$$
+\begin{align}
+H_0 &: \mu_{a} = \mu_{s} \\
+H_1 &: \mu_{a} \neq \mu_{s}
+\end{align}
+$$
+
+Thus, we can perform the following Wilcox test to test this hypothesis.
+
+```{r}
+wilcox.test(diabetic_adult$Glu, diabetic_seniors$Glu, alternative="two.sided")
+```
+
+**TODO :: Figure out this error...**
+
+### 1.3 Vigorous work activity mean blood glucose for adult diabetics
+
+**Task description:** Is it possible to test for the adults diabetic respondents if the mean level of blood glucose after fasting is the same for those who have a vigorous work activity and for those who have not?
+
+Again we will first filter the dataset and create a new combined dataset for these two groups.
+
+```{r}
+diabetic_active <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 1)
+diabetic_inactive <- data %>% filter(Diabete == 1 & age_group == "Adult" & Phys_activ == 2)
+
+diabetic_active$Group <- "Diabetic Active"
+diabetic_inactive$Group <- "Diabetic Inactive"
+
+rmarkdown::paged_table(diabetic_active)
+rmarkdown::paged_table(diabetic_inactive)
+
+# Combine the data frames
+combined_data <- rbind(diabetic_active, diabetic_inactive)
+rmarkdown::paged_table(combined_data)
+```
+
+Immediately we can observe that we have significantly less data than for the other exercises, which can indicate that test results can be inconclusive.
+
+### 1.4 Same proportion of male and female diabetics
+
+**Task description:** Test if the proportion of diabetic is the same for male and female respondents.
+Compare conclusions of several tests.
+You can use the function `table` to compute contingency tables.
+
+## 2. Gene expression data
+
+## 3. Non parametric regression
+
+## 4. S&P500 daily return
+
+### 4.1 Data loading and computing daily return
+
+**Task description:** Load the dataset `sp500_history.csv` into a *data frame* and add a column that computes the daily return for each days.
+
+```{r}
+sp500_data <- read_csv("sp500_history.csv")
+sp500_data$DailyReturn <- sp500_data$Close - sp500_data$Open
+
+# Check for non-numeric values in DailyReturn column
+non_numeric_count <- sum(is.na(sp500_data$DailyReturn) | !is.numeric(sp500_data$DailyReturn))
+cat("Number of non-numeric values in DailyReturn column:", non_numeric_count, "\n")
+
+if (non_numeric_count > 0) {
+  cat("Indices of non-numeric values in DailyReturn column:", which(!is.numeric(sp500_data$DailyReturn)), "\n")
+} else {
+  cat("No non-numeric values found in DailyReturn column.\n")
+}
+
+rmarkdown::paged_table(sp500_data)
+```
+
+### 4.2 Daily return sampled from a normal distribution
+
+**Task description:** We propose to model the daily return as a sample from a normal population. Write the model and use *R* to fit it to the data. What do you think of this model?
+
+To first understand the data, we plot the daily return as a function of the day number.
+
+```{r}
+ggplot(sp500_data, aes(x = Date, y = DailyReturn)) +
+geom_histogram(stat = "identity", fill = "grey", color = "grey", bins = 15) +
+labs(title = "Daily Return of S&P 500 (2007-2010)",
+     x = "Date",
+     y = "Daily Return") +
+theme_minimal() +
+theme(axis.text.x = element_blank(),
+      axis.title.x = element_blank())
+```
+
+Under Louis Bachelier's "Theory of Speculation" assume that the returns can be sampled from a perfectly random distribution, in this case this would be a normal distribution which we can parametirise through a Gaussian model.
+Assume that the return is modeled by some random variable $X_r$ and that some value of that RV is denoted by $x_t$.
+Thus, since we first assume Bachelier's theory we can model the RV through the following Gaussian.
+
+$$    X_r \sim \mathcal{N}(\mu_r, \sigma_r^2)    $$ Where the model parameters are the empirical mean and empirical variance expressed by: $$    \mu_r = \dfrac{\sum_{i=0}^N x_i}{N} \hspace{5mm} \sigma_r^2 = \dfrac{\sum_{i=0}^N (x_i - \mu_r)}{N}    $$ We can then directly fit such a Gaussian model in R through the following:
+
+```{r}
+fit <- fitdistr(sp500_data$DailyReturn, "normal")
+fit
+
+# Plot the histogram of daily returns with the fitted Gaussian distribution
+hist(sp500_data$DailyReturn, breaks = 100, freq = FALSE, 
+     main = "Daily Returns with Fitted Gaussian Distribution",
+     xlab = "Daily Return")
+curve(dnorm(x, mean = fit$estimate[1], sd = fit$estimate[2]), 
+      col = "blue", lwd = 2, add = TRUE, yaxt = "n")
+```
+
+Indeed, the it seems that Bachelier's theory does describe the price movements well.
+As the daily return resembles a normal distribution, however, we do observe that that tails seem to follow a higher variance compared to the density around the mean which seems to have a smaller variance.
+Which also explains why our initial Gaussian doesn't effectively capture this density.
+Thus, we can motivate why a Gaussian mixture model might perform better in this scenario.
+
+### 4.3 Gaussian Mixture Model
+
+**Task description:** Instead of a single normal, we now propose to use a mixture of normals with $p$ components.
+Write the corresponding models and use *R* to fit them to the data for all values of $p$ between $2$ and $6$.
+Discuss the results and compare them with the model of Question 1.
+
+To combine the multiple Gaussians to model a single valid PDF, we will have to introduce a weight parameter which will be used between the weighted sum of Gaussian components.
+First, we denote a single component the following way.
+$$    Y_r \sim \mathcal{N}(\theta_k) \hspace{5mm} \text{Where $\theta$ is a parameter vector for component $k$ defined by} \hspace{5mm} \theta_k = \left[ \mu_k, \sigma_k^2 \right]    $$ Furthermore, to form the mixture we introduce the weight parameter $\pi_k$ which is probabilistic and forms the weighted sum (GMM).
+$$    Y_i = \sum_{k=1}^K \pi_k \mathcal{N}(\theta_k)    $$ As this is a weighted sum and we expect $Y_i$ to be modeled from a valid PDF, we have to constrain the weights such that: $$    \pi_k \geq 0 \hspace{2.5mm} \land \hspace{2.5mm} \sum_{k=1}^K \pi_k = 1    $$ Which is the case as long as we follow that $\pi_k$ comes from a PDF defined over the categorical RVs $Z_i$ which defines the different components such that we have: $$    P(Z_i = k) = \pi_k \hspace{5mm} \text{Where $Z$ is a sequence of RVs} \hspace{5mm} \{1, 2, \dots, K\}    $$ Thus, we observe that we first have to initialize the parameters of the model which we can do by placing the means based on a K-Means initialization and set some explicit variance which will serve as the initialization for the iterative EM-algorithm for parameter optimization over such a GMM.
+First off, we will naively apply a GMM routine to this problem using expectation maximization which we will compare against an constrained optimization approach where we ensure equal means for all distributions as it seems to only be the variance that's needed to change based on our earlier graph.
+
+#### 4.3.1 Unconstrained EM Optimization of GMMs
+
+For this set of experiments we will be using the general GMM approach outlines above, where we initialize the means using k-Means and perform EM to optimize the model parameters.
+
+```{r}
+n_components = 3
+y = sp500_data$DailyReturn
+
+kmeans_out <- kmeans(y, centers = n_components)
+mu = as.vector(kmeans_out$centers)
+mu
+```
+
+This results seems strange given that we should want Gaussian components around the same mean.
+However, for this experiment we will naively apply k-Means initialization without any constraints in the optimization and observe what results we obtain.
+
+```{r}
+# Function which fits n gaussian components and plots the components with the final mixture density.
+fit_mixture_model <- function(n_components, y, max_iter = 400) {
+  kmeans_out <- kmeans(y, centers = n_components)
+  mu <- as.vector(kmeans_out$centers)
+  mixture_EM <- normalmixEM(y, mu = mu, k = n_components, arbvar = FALSE, maxit = max_iter)
+
+  log_likelihood <- mixture_EM$loglik
+  print(log_likelihood)
+  result <- list(pi = mixture_EM$lambda, mu = mixture_EM$mu, sigma = mixture_EM$sigma)
+
+  result <- list(pi = mixture_EM$lambda, mu = mixture_EM$mu, sigma = mixture_EM$sigma)
+
+  # Plot histogram with 50 bars
+  hist(y, breaks = 50, probability = TRUE, col = alpha("skyblue", 0.25), border = alpha("black", 0.25), main = paste("Mixture Model with", n_components, "Components"), xlab = "Daily Return", ylab = "Density")
+
+  # Add density line
+  lines(density(mixture_EM$x), col="red", lwd=2)
+
+  # Plot densities of Gaussian components
+  x <- seq(min(y), max(y), length.out = 1000)
+  for (i in 1:n_components) {
+    curve(result$pi[i] * dnorm(x, mean = result$mu[i], sd = result$sigma[i]), add = TRUE, col = i + 1, lwd=2, lty=2)
+  }
+
+  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood))
+}
+```
+
+Using the above function, we can plot the resulting fits for $2$ to $6$ Gaussian components.
+
+```{r}
+mx2 <- fit_mixture_model(2, y)
+mx3 <- fit_mixture_model(3, y)
+mx4 <- fit_mixture_model(4, y)
+mx5 <- fit_mixture_model(5, y)
+mx6 <- fit_mixture_model(6, y)
+
+mixtures = c(mx2, mx3, mx4, mx5, mx6)
+```
+
+We can observe that the final fits are indeed much better than the single Gaussian example.
+However, we do observe that there is some overfitting happening at the tails of the distribution for all of the experiments we have run.
+This is caused by the fact that the means are not the same for all the Gaussian components and thus for our next section we will be looking at constrained optimization of this GMM where the means should be the same.
+
+#### 4.3.2 Constrained EM Optimisation of GMMs
+
+```{r}
+dcomponents <- function(theta, x) {
+    mapply(
+      function(pi, mu, sigma) pi * dnorm(x, mu, sigma),
+      theta$pi, theta$mu, theta$sigma,
+      SIMPLIFY = TRUE
+    )
+}
+```
+
+```{r}
+# M-step functions, we include the general case such that we can verify the correctness of our implementation.
+
+M_step_general <- function(tau, x) {
+  pi    <- colMeans(tau)
+  mu    <- colSums(tau * x) / colSums(tau)
+  sigma <- sqrt(colSums(tau * x^2) / colSums(tau) - mu^2)
+  list(pi = pi, mu = mu, sigma = sigma)
+}
+
+M_step_same_means <- function(tau, x) {
+  n <- length(x)
+  K <- ncol(tau)
+  pi    <- colSums(tau) / n
+  muk   <- colSums(tau * x) / colSums(tau)
+  sigma <- sqrt(colSums(tau * x^2) / colSums(tau) - muk^2)
+  wk    <- n * pi / sigma^2
+  mu    <- rep(sum(wk * muk) / sum(wk), K)
+  list(pi = pi, mu = mu, sigma = sigma)
+}
+```
+
+```{r}
+mixture_gaussian1D <- function(x, theta0, M_step = M_step_general, max_iter = 100, threshold = 1e-6) {
+
+  ## initialization
+  n <- length(x)
+  likelihoods  <- dcomponents(theta0, x) 
+  deviance     <- numeric(max_iter)
+  deviance[1]  <- -2 * sum(log(rowSums(likelihoods)))
+
+  for (t in 1:max_iter) {
+
+    # E step
+    tau <- likelihoods / rowSums(likelihoods)
+    # M step
+    theta <- M_step(tau, x)
+
+    ## Assessing convergence
+    likelihoods   <- dcomponents(theta, x)
+    deviance[t+1] <- - 2 * sum(log(rowSums(likelihoods)))
+
+    ## prepare next iterations
+
+    if (abs(deviance[t + 1] - deviance[t]) < threshold)
+      break
+
+  }
+
+  list(theta = theta, deviance = deviance[t + 1])
+}
+```
+
+```{r}
+mu_data = mean(y)
+pi = c(.5, .5)
+means = c(mu_data, mu_data)
+variances = c(10,2)
+
+theta0  <- list(pi=pi, mu=means, sigma=variances)
+out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)
+out_general$theta
+```
+
+To then compare the results of using different number of components we additionally define a experiment function.
+
+```{r}
+softmax <- function(x) {
+  exp_x <- exp(x)
+  exp_x / sum(exp_x)
+}
+
+constrained_gmm_experiment <- function(y, variances) {
+  n_components = length(variances)
+
+  mu_data = mean(y)
+
+  pi = softmax(rep(1, n_components))
+  means = rep(mu_data, n_components)
+
+  theta0  <- list(pi=pi, mu=means, sigma=variances)
+  out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)
+  out_general$theta
+
+  # Define the data
+  x <- seq(-150, 150, length.out = 1000)
+
+  # Compute the densities for the two components
+  densities <- dcomponents(out_general$theta, x)
+
+  # Calculate the final density of the mixture
+  final_density <- rowSums(sapply(1:length(out_general$theta$pi), function(i) out_general$theta$pi[i] * dnorm(x, mean = out_general$theta$mu[i], sd = out_general$theta$sigma[i])))
+
+  # Plot the data points
+  hist(y, breaks = 50, probability = TRUE, col = alpha("skyblue", 0.25), border = alpha("black", 0.25), main = paste("Mixture Model with", n_components, "Components"), xlab = "Daily Return", ylab = "Density")
+
+  # Plot the densities for each component
+  for (i in 1:n_components) {
+    lines(x, densities[,i], col = rainbow(n_components)[i], lwd=2, lty=2)
+  }
+
+  # Plot the final density of the mixture
+  lines(x, final_density, lwd = 2)
+
+  # Add legend
+  legend_labels <- c(paste0("Component ", 1:n_components), "Mixture")
+  legend("topright", legend = legend_labels, col = c(rainbow(n_components), "black"), lty = 1, lwd = 2, pch = 20)
+}
+```
+
+```{r}
+constrained_gmm_experiment(y, variances = c(10, 5))
+```
+
+### 4.4 Student t-distribution
+
+**Task description:** As a third and last model, we propose to model the daily return as coming from a location-scale family of Student distributions, which is a model defined by the collection of densities $$f_{\nu,m,a}(x) = \frac{\Gamma\big(\frac{\nu+1}{2}\big)}{\sqrt{\pi \nu a^2}\Gamma\big(\frac{\nu}{2}\big)}\Bigg(1+\frac{(x-m)^2}{\nu a^2} \Bigg)^{-\frac{\nu+1}{2}}$$ where $\Gamma$ is the Gamma function and $\theta=(\nu,m,a) \in \mathbb{R}_+^*\times \mathbb{R}\times \mathbb{R}_+^*$ is the parameter.
+
+#### 4.4.1 Motivation behind model selection
+
+**Task description:** What is the motivation for considering this model?
+
+Indeed from looking at the above examples the final fits are still not great even when we add additional components to account for the variable variance of the distribution.
+But as we observe the distribution it is more clear that it can follow more closely a student t-distribution given its long tails and elongated middle density.
+
+#### 4.4.2 Fitting the student t-distribution
+
+**Task description:** Propose an algorithm for estimating the parameter of this model and use it to fit this model to the data.
+
+```{r}
+# Define log-likelihood function for Student's t-distribution
+log_likelihood_student_t <- function(theta, x) {
+  nu <- theta[1]
+  mu <- theta[2]
+  sigma <- theta[3]
+
+  # Check if sigma is positive
+  if(sigma <= 0) {
+    return(Inf)
+  }
+
+  # Check if nu is at least 1
+  if(nu < 1) {
+    nu <- 1
+  }
+
+  loglik <- sum(dt((x - mu) / sigma, df = nu, log = TRUE)) - length(x) * log(sigma)
+
+  return(-loglik)  # Negative log-likelihood for minimization
+}
+
+# Fit Student's t-distribution model using MLE
+fit_student_t <- function(y) {
+  if (any(!is.finite(y))) {
+    stop("Input data contains non-finite values.")
+  }
+
+  y <- y[is.finite(y)]
+
+  if (length(y) < 3) {
+    stop("Insufficient data points for estimation.")
+  }
+
+  init_params <- c(5, mean(y), sd(y))
+  opt_result <- optim(init_params, log_likelihood_student_t, x = y, method = "L-BFGS-B",
+                      lower = c(0, -Inf, 0), upper = c(Inf, Inf, Inf))
+
+  # Extract optimized parameters
+  nu <- opt_result$par[1]
+  mu <- opt_result$par[2]
+  sigma <- opt_result$par[3]
+
+  # Return parameter estimates
+  return(list(nu = nu, mu = mu, sigma = sigma))
+}
+
+# Fit Student's t-distribution model to the data
+student_t_params <- fit_student_t(y)
+student_t_params
+```
+
+```{r}
+# Generate data for the PDF using the fitted parameters
+x_range <- seq(min(y), max(y), length.out = 1000)
+pdf_values <- dt((x_range - student_t_params$mu) / student_t_params$sigma, 
+                 df = student_t_params$nu) / student_t_params$sigma
+
+# Create a data frame for plotting
+plot_data <- data.frame(x = x_range, pdf = pdf_values)
+
+# Plot histogram of the data and fitted PDF
+ggplot() +
+  geom_histogram(data = data.frame(y), aes(x = y, y = ..density..), bins = 30, fill = "skyblue", color = "black") +
+  geom_line(data = plot_data, aes(x = x, y = pdf), color = "red", size = 1) +
+  labs(title = "Fitted Student's t-distribution Model",
+       x = "Daily Return",
+       y = "Density") +
+  theme_minimal()
+```

### 2024-03-04 submission.Rmd
chore: added subsections i forgot
feat: e-step explained (m-step remaining)

@@
 
+To be able to optimize a GMM where the component means are constrained to be equal, we have to modify the classic expectation maximization algorithm to account for this specific constraint.
+First off, lets define the basic function for computing the density of a GMM which we will use later.
+
 ```{r}
@@
 ```
+
+Now that we have the above definition of a GMM density we can define how our new EM algorithm is going to work.
+As the EM algorithm is broken into two steps (E-step and M-step), we will first define the E-step.
+
+**E-step**
+
+As presented earlier we assume that we have some labels $Z_i$ which indicate what components model what sub-parts of the data.
+However, this is not something that we can know from beforehand.
+Thus, we compute the expected value of the statistic $\mathbb{E}[S(Z, Y) | Y ; \theta^{(t-1)}]$.
+Which we can denote using $\tau_{ik}^t$.
+
+$$
+\begin{align}
+\tau_{ik}^t &= \mathbb{E}[S(Z,Y) | Y; \theta^{(t-1)}] \\
+            &= \dfrac{\pi_k^{(t-1)} f_k(y_i;\theta^{(t-1)})}{\sum_{\mathcal{l}=0}^K \pi_k^{(t-1)} f_{\mathcal{l}} (y_i ; \theta^{(t-1)})}
+\end{align}
+$$
+
+Indeed, we see that there is no need to change anything in our E-step to achieve this constraint. However, we will have to deal with this in the M-step.
+
+**M-step**
 
@@
 ```
+
+### 4.5 Model evaluation
+
+**Task description:** Between all the previous models (the normal, the 5 mixtures of normals, and the location-scale Student) which one do you choose? Explain your methodology.
+
+### 4.6 Expected daily return
+
+**Task description:** Is the expected daily return different than zero?

### 2024-03-04 submission.Rmd
feat: m-step + evaluation

@@
 
-  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood))
+  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood, deviance= ))
 }
@@
 **M-step**
+
+For the M-step we follow a similar procedure but now we have to change the update for the means. Which will take the following form. First we have the same parameters,
+
+$$
+\begin{align}
+\hat{\pi}_k = \dfrac{\sum_{i=1}^K \tau_{ik}}{n} \hspace{5mm} \mu_k = \dfrac{\sum_{i=1}^K \tau_{ik} x_i}{\sum_{i=1}^K \tau_{ik}} \hspace{5mm} \hat{\sigma}_k^2 = \dfrac{\sum_{i=1}^K \tau_{ik} z_i^2}{\sum_{i=1}^K \tau_{ik}} - \mu_k^2
+\end{align}
+$$
+
+Note that we do not denote the mean with $\hat{\mu}$ above as it is only used on the optimization for the variance, to compute the new $\hat{\mu}$ we do the following.
+
+$$
+\hat{\mu}_k = \frac{\sum_{i=1}^K w_k \mu_k}{\sum_{i=1}^K w_k}
+$$
+
+Where $w_k$ is computed by.
+
+$$
+w_k = \dfrac{\sum_{i=1}^K \tau_{ik}}{\hat{\sigma}_{k}^2}
+$$
+
+We can then use this to create a modified version of the EM algorithm to deterministically obtain components with equal means.
 
@@
 # Define log-likelihood function for Student's t-distribution
-log_likelihood_student_t <- function(theta, x) {
+neglog_likelihood_student_t <- function(theta, x) {
   nu <- theta[1]
@@
   init_params <- c(5, mean(y), sd(y))
-  opt_result <- optim(init_params, log_likelihood_student_t, x = y, method = "L-BFGS-B",
+  opt_result <- optim(init_params, neglog_likelihood_student_t, x = y, method = "L-BFGS-B",
                       lower = c(0, -Inf, 0), upper = c(Inf, Inf, Inf))
@@
 ggplot() +
-  geom_histogram(data = data.frame(y), aes(x = y, y = ..density..), bins = 30, fill = "skyblue", color = "black") +
+  geom_histogram(data = data.frame(y), aes(x = y, y = after_stat(density)), bins = 30, fill = "skyblue", color = "black") +
   geom_line(data = plot_data, aes(x = x, y = pdf), color = "red", size = 1) +
@@
 
+To effectively evaluate the models we can employ a few techniques, the most obvious is to apply the BIC metric to measure how well the models capture the distribution of the data.
+
+```{r}
+bic_benchmark <- function(x, out_model, n_free_params) {
+  n <- length(x)
+  bic_values = list()
+
+  return(out_model$deviance + log(n) * n_free_params)
+}
+```
+
+```{r}
+mu_data = mean(y)
+pi = c(.2, .2, .2, .2, .2)
+means = c(mu_data, mu_data, mu_data, mu_data, mu_data)
+variances = c(10, 7, 5, 3, 1)
+
+theta0  <- list(pi=pi, mu=means, sigma=variances)
+out_general <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_general)
+
+b_5unconstrained <- bic_benchmark(y, out_general, 3 * 5)
+
+out_general_constrained <- mixture_gaussian1D(sp500_data$DailyReturn, theta0, M_step = M_step_same_means)
+
+b_5constrained <- bic_benchmark(y, out_general_constrained, 3 * 5)
+
+theta = student_t_params
+b_tdist = 3 * log(length(y)) + 2 * neglog_likelihood_student_t(c(theta$nu, theta$mu, theta$sigma), y)
+
+single_guass <- fitdistr(sp500_data$DailyReturn, "normal")
+b_single_gauss <- BIC(single_guass)
+```
+
+```{r}
+rbind(
+  five_comp_gmm             = b_5unconstrained,
+  five_comp_gmm_constrained = b_5constrained,
+  student_tdistribution     = b_tdist,
+  single_gaussian           = b_single_gauss
+) %>% as.data.frame() %>% setNames("BIC") %>% rmarkdown::paged_table()
+```
+
+```{r}
+ks.test(y, "pmixture", out_general_constrained$theta)
+```
+
+```{r}
+ks.test(y, "pmixture", out_general$theta)
+```
+
+```{r}
+ks.test(y, "pmixture", theta0)
+```
+
 ### 4.6 Expected daily return

### 2024-03-05 submission.Rmd
feat: model evaluation done

@@
   mixture_EM <- normalmixEM(y, mu = mu, k = n_components, arbvar = FALSE, maxit = max_iter)
-
   log_likelihood <- mixture_EM$loglik
-  print(log_likelihood)
-  result <- list(pi = mixture_EM$lambda, mu = mixture_EM$mu, sigma = mixture_EM$sigma)
 
@@
   # Add density line
-  lines(density(mixture_EM$x), col="red", lwd=2)
+  final_density <- rowSums(sapply(1:length(mixture_EM$lambda), function(i) mixture_EM$lambda[i] * dnorm(x, mean = mixture_EM$mu[i], sd = mixture_EM$sigma[i])))
+
+  x <- seq(min(y), max(y), length.out = 1000)
+  lines(x, final_density, lwd = 2)
 
   # Plot densities of Gaussian components
-  x <- seq(min(y), max(y), length.out = 1000)
   for (i in 1:n_components) {
@@
 
-  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood, deviance= ))
+  return(list(result = result, mixture_EM = mixture_EM, log_likelihood = log_likelihood))
 }
@@
   # Define the data
-  x <- seq(-150, 150, length.out = 1000)
+  x <- seq(min(y), max(y), length.out = 1000)
 
@@
 ```{r}
-constrained_gmm_experiment(y, variances = c(10, 5))
+constrained_gmm_experiment(y, variances = c(10, 7))
+constrained_gmm_experiment(y, variances = c(10, 7, 5))
+constrained_gmm_experiment(y, variances = c(10, 7, 5, 3))
+constrained_gmm_experiment(y, variances = c(10, 7, 5, 3, 1))
+constrained_gmm_experiment(y, variances = c(12, 10, 7, 5, 3, 1))
 ```
@@
 To effectively evaluate the models we can employ a few techniques, the most obvious is to apply the BIC metric to measure how well the models capture the distribution of the data.
+Which is simply done by applying the following formula:
+
+$$
+BIC = k \hspace{1mm} ln(n) - 2 \hspace{1mm} ln(\hat{L})
+$$
+
+Where we have the following parameters to compute:
+
+-   $\hat{L}$: the maximized likelihood of the model.
+
+-   $n$: the number of data points in the dataset.
+
+-   $k$: the number of free parameters in our model (note for the constrained GMM this will be different).
+
+We then go ahead and apply this metric to the models so far and put the results in a table seen below.
 
@@
 
-b_5constrained <- bic_benchmark(y, out_general_constrained, 3 * 5)
+b_5constrained <- bic_benchmark(y, out_general_constrained, 2 * 5 + 1) # Only one mu is optimised
 
@@
 
+By only observing the BIC values it might seem that the single Gaussian is the best model.
+However, this should not be the case, indeed we need to perform additional tests.
+The Kolmogorov-Smirnov test can work for this, and we can expect to see that the single Gaussian is rejected.
+Where we have the following null- and alternative hypothesis
+
+$$
+H_0: \text{the empirical distribution of the data fits the theoretical distribution} \\
+H_1: \text{The empirical distribution does not fit the theoretical distribution}
+$$
+
 ```{r}
@@
 
+```{r}
+# Function to compute CDF of Student's t-distribution
+cdf_student_t <- function(x, nu, mu, sigma) {
+  pt((x - mu) / sigma, df = nu)
+}
+
+# Perform KS test
+ks_test_student_t <- function(y, student_t_params) {
+  # Extract parameters
+  nu <- student_t_params$nu
+  mu <- student_t_params$mu
+  sigma <- student_t_params$sigma
+
+  # Compute CDF values
+  cdf_values <- cdf_student_t(sort(y), nu, mu, sigma)
+
+  # Compute ECDF values
+  ecdf_values <- ecdf(y)(sort(y))
+
+  # Perform KS test
+  ks_result <- ks.test(ecdf_values, cdf_values)
+
+  return(ks_result)
+}
+
+# Perform KS test on the data using the fitted Student's t-distribution parameters
+ks_result <- ks_test_student_t(y, student_t_params)
+ks_result
+
+```
+
+Thus, we can quite confidently conclude that we can reject the single Gaussian as a model (as the null-hypothesis is rejected) and announce the Student t-distribution to be the best distribution given the data.
+Which makes intuitive sense given the fact that the distribution of the data much more resemble that of a Student t-distribution.
+
 ### 4.6 Expected daily return

### 2024-03-08 submission.Rmd
Max added q2

@@
 ## 2. Gene expression data
+
+```{r}
+load("liver_data.rda")
+```
+
+### 2.1  We first want to study the gene which is the most correlated with the level of cholesterol.
+Identify this gene and plot `cholesterol` as a function of this gene expression level.
+Fit a linear model.
+Test if there is a significant relation between this gene expression and `cholesterol`?
+
+```{r}
+labels <- colnames(liver)
+
+maxSlope=0
+minP=1
+
+for (i in 2:3117){
+  lmgene <- lm(cholesterol~liver[,i], data=liver)
+  slope <- abs(coef(lmgene, complete = TRUE)[2])
+  summary_stats <- summary(lmgene)
+  p_value <- summary_stats$coefficients[2, 4]
+
+  if (slope > maxSlope){
+    maxSlope=slope
+    max_slope_model=lmgene
+  }
+
+  if (minP > p_value){
+    minP=p_value
+    min_p_model=lmgene
+    genename=labels[i]
+  }
+}
+
+```
+
+Fitting a linear model for every gene, we get 3116 different models in total.
+The null-hypothesis is that a gene does no have a correlation with the cholesterol-level, in other words, the slope is zero in this case.
+Consequently, the gene corresponding to the model with the lowest p-value is also the model with highest slope.
+In our case, that turned out to be gene $A\_42P631473$ with a p-value of less than $10^{-13}$.
+
+```{r}
+
+ggplotRegression <- function (fit) {
+
+require(ggplot2)
+
+ggplot(fit$model, aes_string(x = names(fit$model)[2], y = names(fit$model)[1])) + 
+  geom_point() +
+  stat_smooth(method = "lm", col = "red") +
+  labs(title = paste("Adj R2 = ",signif(summary(fit)$adj.r.squared, 5),
+                     "Intercept =",signif(fit$coef[[1]],5 ),
+                     " Slope =",signif(fit$coef[[2]], 5),
+                     " P =",signif(summary(fit)$coef[2,4], 5)), x = paste(genename, "-expression"))
+}
+
+ggplotRegression(min_p_model)
+#ggplotRegression(max_slope_model)
+
+```
+
+## 2.2  We now wish to perform the previous test for all of the 3116 genes.
+For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
+AND
+
+## 2.3  Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+
+```{r}
+labels <- colnames(liver)
+
+p_list <- c()
+lineList = c()
+
+for (i in 2:3117){
+  lmgene <- lm(cholesterol~liver[,i], data=liver)
+  slope <- abs(coef(lmgene, complete = TRUE)[2])
+  summary_stats <- summary(lmgene)
+  p_value <- summary_stats$coefficients[2, 4]
+  p_list=c(p_list, p_value)
+  lineList = c(lineList, (i-1)/3116)
+}
+
+p_list=sort(p_list)
+plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
+lines(lineList, col = "red", lwd = 2)
+
+legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
+
+```
+Plotting all values, we can conclude that there are amount of models that have a p-value below 0.05.
+It is tempting to make the conclusion that all these models are statistically significant since they all have a small p-value.
+However, in the event that the we only have random data, we still expect 5\% of the models to appear significantly significant.
+In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
+Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+
+```{r}
+labels <- colnames(liver)
+
+p_list_shuffeled <- c()
+lineList_shuffeled = c()
+
+for (i in 2:3117){
+  cholesterol=liver$cholesterol
+  randomLabel <- sample(cholesterol)
+  lmgene <- lm(randomLabel~liver[,i], data=liver)
+  slope <- abs(coef(lmgene, complete = TRUE)[2])
+  summary_stats <- summary(lmgene)
+  p_value <- summary_stats$coefficients[2, 4]
+  p_list_shuffeled=c(p_list_shuffeled, p_value)
+  lineList_shuffeled = c(lineList_shuffeled, (i-1)/3116)
+}
+
+p_list_shuffeled=sort(p_list_shuffeled)
+plot(p_list_shuffeled, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value", main = "Shuffeled labels")
+lines(lineList_shuffeled, col = "red", lwd = 2)
+
+legend("topleft", legend = c("p-values when shuffeled labels", "Uniform line"), col = c("blue", "red"), lwd = 2)
+
+```
+ In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5\%.
+For this method, we use the Benjamini-Hochberg procedure \cite{FDR}.
+In short, it states that if we have a set of $\{ H_1, H_2, ...
+\: , H_m \}$ with associated and ordered p-values $\{ p_1, p_2, ...
+\: , p_m \}$, then the threshold for the index can be written as:
+
+##2.4
+
+$$ P_{BH}= max_i \left\{ P_{(i)} : P_{(i)} \leq \alpha \frac{i}{m}      \right\} $$
+
+```{r}
+
+soft_guarantee=c()
+alpha=0.05
+index=0
+n=3116
+
+for (i in 1:3116){
+  val=i*alpha/n
+  soft_guarantee=c(soft_guarantee, val)
+  if (p_list[i]<val){
+    index=i
+  }
+
+}
+
+print(index)
+
+plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
+lines(soft_guarantee, col = "red", lwd = 2)
+abline(v=index, col="black")
+legend("topleft", legend = c("Benjamini-Hochberg boundary", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)
+
+plot(p_list[1:1200], col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
+lines(soft_guarantee[1:1200], col = "red", lwd = 2)
+abline(v=index, col="black")
+legend("topleft", legend = c("Benjamini-Hochberg boundary", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)
+
+```
+We can see that $\alpha \frac{i}{m}$ plots out a line, and we choose i such that $P_{(i)}$ always is smaller or equal to that value.
+Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 \%
+
+##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 \% risk that any discovery is false.
+For this method, we use the Holm–Bonferroni method.
+The method is as follows:\\
+
+I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+Is $P_1 < \alpha/m$? so, reject $H_1$ and continue to the next step, otherwise EXIT.
+Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
+
+Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$. Using this process, we get the following result:
+
+```{r}
+
+strong_guarantee=c()
+alpha=0.05
+index=0
+
+for (i in 1:n){
+  val=alpha/(n+1-i)
+  strong_guarantee=c(strong_guarantee, val)
+  if (p_list[i]<val){
+    index=i
+  }
+
+}
+
+plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
+lines(strong_guarantee, col = "red", lwd = 2)
+abline(v=index, col="black", )
+legend("topleft", legend = c("Holm–Bonferroni", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)
+
+plot(p_list[1:225], col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
+lines(strong_guarantee, col = "red", lwd = 2)
+abline(v=index, col="black", )
+legend("bottomright", legend = c("Holm–Bonferroni", "P-value", paste("Intersect=", as.character(index))), col = c("red", "blue", "black"), lwd = 2)
+
+```
+
+Using this process, we see that we get 223 discoveries. We also not that smallest p-value in that set is approximately equal to $1.7 \cdot 10^{-5}$.
 

### 2024-03-08 submission.Rmd
feat: general improvements

--- 2024-03-05 submission.Rmd
@@
 library(mixtools)
+library(pwr)
 ```
@@
 
+Furthermore, it is worth noting that the sample size between the two groups are different and thus the traditional t-test mechanisms are changed to account for this.
+However, the techniques used above will be able to handle this internally.
+
 ### 1.2 Diabetic mean blood glucose level between adults and seniors
@@
 Immediately we can observe that we have significantly less data than for the other exercises, which can indicate that test results can be inconclusive.
+To verify how many samples we should have to be able to be able to perform such a test, we can use the power of a t-test.
+To apply this test we have to understand what
+
+```{r}
+
+# Perform power analysis
+n_active <- 4  # Sample size of the active group
+n_inactive <- 9  # Sample size of the inactive group
+
+sp <- sqrt(((n_active-1)*sd(diabetic_active$Glu)**2 + (n_inactive - 1)*sd(diabetic_inactive$Glu)**2**2)/(n_active+n_inactive-2))
+effect_size <- (mean(diabetic_active$Glu) - mean(diabetic_inactive$Glu))/sp
+alpha <- 0.05  # Significance level
+
+# Perform power analysis
+pwr.t.test(n = NULL, d = effect_size, sig.level = alpha, power = 0.5,
+                              alternative = "two.sided")
+```
 

### 2024-03-09 submission.Rmd
feat: ex1.3 power.t.test we need to figure this one out

--- 2024-03-08 submission.Rmd
@@
 library(mixtools)
+library(pwr)
+library(mclust)
 ```
@@
 ```{r}
-wilcox.test(diabetic_adult$Glu, diabetic_seniors$Glu, alternative="two.sided")
+delta_wilcox <- wilcox.test(diabetic_seniors$Glu, diabetic_adult$Glu, alternative="two.sided")
+delta_wilcox
 ```
 
-**TODO :: Figure out this error...**
+Based on these results we can observe that we cannot reject the null-hypothesis, thus, we can say that there is not enough evidence to suggest that the mean blood glucose between diabetic adults and diabetic seniors is different.
 
@@
 rmarkdown::paged_table(combined_data)
+
+ggplot(combined_data, aes(x = Group, y = Glu)) +
+  geom_boxplot() +
+  labs(x = "Activity Group", y = "Glucose") +
+  ggtitle("Glucose Distribution for Diabetics by Activity") +
+  theme_minimal()
+
+ggplot(combined_data, aes(x = Group, y = Glu, fill = Group)) +
+  geom_violin(trim = FALSE) +
+  labs(x = "Activity Group", y = "Glucose") +
+  ggtitle("Glucose Distribution for Diabetics by Activity") +
+  theme_minimal()
 ```
@@
 Immediately we can observe that we have significantly less data than for the other exercises, which can indicate that test results can be inconclusive.
+Furthermore, we cannot necessarily make an assumption about the normality of the data due to the fact that we have such little samples.
+It could for example be the case that for many more samples that the distribution looks much more like a normal.
+
+However, for now we will not make any normality assumption and we will be using a Wilcoxon test where we will analyse its power of varying number of samples for $\alpha=0.05$ which will enable us to see how the power of the test evolves over the number of samples.
+
+```{r}
+delta_wilcox  <- wilcox.test(diabetic_active$Glu, diabetic_inactive$Glu, alternative="two.sided")$statistic
+alpha <- 0.05
+seq_n <- seq(10,1)
+power_wilcox  <- map_dbl(seq_n, ~power.t.test(n = .x, delta_wilcox , sig.level = alpha)$power)
+data.frame(power = c(power_wilcox), n = rep(seq_n),
+           method = rep(c("Wilcox"), each = length(seq_n))) %>% 
+  ggplot() + aes(x = n, y = power, color = method) + geom_line()
+```
+
+We observe that for a sample size of 4, the power of the Wilcoxon t-test is more or less 1.0 meaning that it is almost guaranteed that the null-hypothesis will be rejected when it is false.
+Furthermore, we plot for $n=4$ the different values of $\alpha$ wrt.
+the power of the Wilcoxon test.
+
+```{r}
+delta_wilcox  <- wilcox.test(diabetic_active$Glu, diabetic_inactive$Glu, alternative="two.sided")$statistic
+seq_alpha <- 10^seq(-12, -5, .1)
+n <- 4
+power_wilcox  <- map_dbl(seq_alpha, ~power.t.test(n, delta_wilcox , sig.level = .x)$power)
+data.frame(power = c(power_wilcox), alpha = rep(seq_alpha, 1),
+           method = rep(c("Wilcox"), each = length(seq_alpha))) %>% 
+  ggplot() + aes(x = alpha, y = power, color = method) + geom_line()
+```
+
+As observed earlier for a $\alpha=0.05$ we basically have guaranteed that the null-hypothesis is rejected correctly only after 2 samples.
+However, this seems strange and thus one might want to adjust the $\alpha$ value to get a more reasonable sample requirement.
 
@@
 You can use the function `table` to compute contingency tables.
+
+```{r}
+diabete_and_nondiabete = filter(data, Diabete != 3)
+contingency_table <- table(diabete_and_nondiabete$Sex, diabete_and_nondiabete$Diabete)
+contingency_table
+```
+
+```{r}
+fisher.test(contingency_table)
+```
+
+```{r}
+diabetes = c(8, 13)
+population = c(nrow(filter(data, Sex==1)), nrow(filter(data, Sex==2)))
+
+prop.test(diabetes, population, correct = FALSE)
+```
+
+```{r}
+prop.test(diabetes, population, correct = TRUE)
+```
 
@@
 
-## 2.2  We now wish to perform the previous test for all of the 3116 genes.
+### 2.2 We now wish to perform the previous test for all of the 3116 genes.
 For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
@@
 
-## 2.3  Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+### 2.3 Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
 
@@
 ```
+
 Plotting all values, we can conclude that there are amount of models that have a p-value below 0.05.
 It is tempting to make the conclusion that all these models are statistically significant since they all have a small p-value.
-However, in the event that the we only have random data, we still expect 5\% of the models to appear significantly significant.
+However, in the event that the we only have random data, we still expect 5% of the models to appear significantly significant.
 In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
@@
 ```
- In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5\%.
+
+In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5%.
 For this method, we use the Benjamini-Hochberg procedure \cite{FDR}.
@@
 ```
+
 We can see that $\alpha \frac{i}{m}$ plots out a line, and we choose i such that $P_{(i)}$ always is smaller or equal to that value.
-Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 \%
+Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 %
 
-##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 \% risk that any discovery is false.
+##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false.
 For this method, we use the Holm–Bonferroni method.
@@
 
-I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
-Is $P_1 < \alpha/m$? so, reject $H_1$ and continue to the next step, otherwise EXIT.
-Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
+I have my ordered p-values, $\{ P_1, P_2, ...
+\: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+Is $P_1 < \alpha/m$?
+so, reject $H_1$ and continue to the next step, otherwise EXIT.
+Is $P_2 < \alpha/(m-1)$?
+so, reject $H_2$ and continue to the next step, otherwise EXIT.
 

### 2024-03-09 submission.Rmd
MAde small changes and made it astethically pleasing

--- 2024-03-08 submission.Rmd
@@
 library(mixtools)
-library(pwr)
 ```
@@
 
-Furthermore, it is worth noting that the sample size between the two groups are different and thus the traditional t-test mechanisms are changed to account for this.
-However, the techniques used above will be able to handle this internally.
-
 ### 1.2 Diabetic mean blood glucose level between adults and seniors
@@
 Immediately we can observe that we have significantly less data than for the other exercises, which can indicate that test results can be inconclusive.
-To verify how many samples we should have to be able to be able to perform such a test, we can use the power of a t-test.
-To apply this test we have to understand what
-
-```{r}
-
-# Perform power analysis
-n_active <- 4  # Sample size of the active group
-n_inactive <- 9  # Sample size of the inactive group
-
-sp <- sqrt(((n_active-1)*sd(diabetic_active$Glu)**2 + (n_inactive - 1)*sd(diabetic_inactive$Glu)**2**2)/(n_active+n_inactive-2))
-effect_size <- (mean(diabetic_active$Glu) - mean(diabetic_inactive$Glu))/sp
-alpha <- 0.05  # Significance level
-
-# Perform power analysis
-pwr.t.test(n = NULL, d = effect_size, sig.level = alpha, power = 0.5,
-                              alternative = "two.sided")
-```
 
@@
 
-### 2.1  We first want to study the gene which is the most correlated with the level of cholesterol.
+### 2.1 Finding the most significant Gene
+
+We first want to study the gene which is the most correlated with the level of cholesterol.
 Identify this gene and plot `cholesterol` as a function of this gene expression level.
@@
 
-## 2.2  We now wish to perform the previous test for all of the 3116 genes.
-For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
-AND
+## 2.2 Calculating all p-values
 
-## 2.3  Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+Plotting the p-values of all models, we see that a great number of genes correspond to a p-value of less that 0.05.
+At a first glance, genes as a whole would appear to contain information that could help us make discoveries.
 
@@
 
+```
+
+```{r}
+plot(hist(p_list, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
+```
+
+## 2.3 Ordering p-values
+
+Plotting all values, we can conclude that there are amount of models that have a p-value below 0.05.
+It is tempting to make the conclusion that all these models are statistically significant since they all have a small p-value.
+However, in the event that the we only have random data, we still expect 5% of the models to appear significantly significant.
+In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
+Visually, looking at the Figure below, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+
+```{r}
+
 p_list=sort(p_list)
@@
 ```
-Plotting all values, we can conclude that there are amount of models that have a p-value below 0.05.
-It is tempting to make the conclusion that all these models are statistically significant since they all have a small p-value.
-However, in the event that the we only have random data, we still expect 5\% of the models to appear significantly significant.
-In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
-Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+
+Shuffling the labels, we get the case that the p-values are approximately uniformly distributed and that the p-values follow the line
 
@@
 
+plot(hist(p_list_shuffeled, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
 ```
- In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5\%.
-For this method, we use the Benjamini-Hochberg procedure \cite{FDR}.
+
+## 2.4 Benjamini-Hochberg procedure
+
+In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5%.
+For this method, we use the Benjamini-Hochberg procedure.
 In short, it states that if we have a set of $\{ H_1, H_2, ...
@@
 \: , p_m \}$, then the threshold for the index can be written as:
-
-##2.4
 
@@
 
-print(index)
-
 plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
@@
 ```
+
 We can see that $\alpha \frac{i}{m}$ plots out a line, and we choose i such that $P_{(i)}$ always is smaller or equal to that value.
-Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 \%
+Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 %
 
-##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 \% risk that any discovery is false.
-For this method, we use the Holm–Bonferroni method.
-The method is as follows:\\
+## 2.5 Holm–Bonferroni method
 
-I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
-Is $P_1 < \alpha/m$? so, reject $H_1$ and continue to the next step, otherwise EXIT.
-Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
+A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false. For this method, we use the Holm–Bonferroni method. The method is as follows:\\
+
+I have my ordered p-values, $\{ P_1, P_2, ...
+\: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+Is $P_1 < \alpha/m$?
+so, reject $H_1$ and continue to the next step, otherwise EXIT.
+Is $P_2 < \alpha/(m-1)$?
+so, reject $H_2$ and continue to the next step, otherwise EXIT.
 
@@
 
-Using this process, we see that we get 223 discoveries. We also not that smallest p-value in that set is approximately equal to $1.7 \cdot 10^{-5}$.
+Using this process, we see that we get 223 discoveries. We also not that biggest p-value in that set is approximately equal to $1.7 \cdot 10^{-5}$.
 

### 2024-03-09 submission.Rmd

--- 2024-03-09 submission.Rmd
@@
 
-### 2.2 We now wish to perform the previous test for all of the 3116 genes.
+## 2.2  We now wish to perform the previous test for all of the 3116 genes.
 For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
@@
 
-### 2.3 Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+## 2.3  Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
 
@@
 
-p_list=sort(p_list)
-plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
-lines(lineList, col = "red", lwd = 2)
-
-legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
-
-```
+```
+
+```{r}
+plot(hist(p_list, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
+```
+
+## 2.3 Ordering p-values
 
@@
 In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
-Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+Visually, looking at the Figure below, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+
+```{r}
+
+p_list=sort(p_list)
+plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
+lines(lineList, col = "red", lwd = 2)
+
+legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
+
+```
 
@@
 ```
-
-In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5%.
+ In the nest question, we are tasked with finding the maximum number of genes such that the expected number of false discoveries is at most 5\%.
 For this method, we use the Benjamini-Hochberg procedure \cite{FDR}.
@@
 
-##2.4
+plot(hist(p_list_shuffeled, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
 
@@
 ```
-
 We can see that $\alpha \frac{i}{m}$ plots out a line, and we choose i such that $P_{(i)}$ always is smaller or equal to that value.
-Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 %
-
-##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false.
+Doing so, we get 941 discoveries such that the expected number of false discoveries is no more than 5 \%
+
+##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 \% risk that any discovery is false.
 For this method, we use the Holm–Bonferroni method.

### 2024-03-09 submission.Rmd
updated q2

--- 2024-03-09 submission.Rmd
@@
 
-### 2.1 We first want to study the gene which is the most correlated with the level of cholesterol.
+## 2.1 Finding most significant discovery
+
+We first want to study the gene which is the most correlated with the level of cholesterol.
 Identify this gene and plot `cholesterol` as a function of this gene expression level.
@@
 
-### 2.2 We now wish to perform the previous test for all of the 3116 genes.
-For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
-AND
-
-### 2.3 Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+## 2.2 Calculating all p-values
+
+We now do the same test for all 3116 genes and fit a linear model that explains `cholesterol` as a function of the gene expression.
+Doing so gives us a relatively large proportion of genes that correspond to a p-value of less than 0.05
 
@@
 In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
-Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
+Visually, looking at the figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant discoveries.
+
+```{r}
+
+p_list=sort(p_list)
+plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
+lines(lineList, col = "red", lwd = 2)
+
+legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
+
+```
+
+Shuffling all labels, we get an approximate uniform distribution of p-values which corresponds to the p-values following a straight line
 
@@
 
-```
+plot(hist(p_list_shuffeled, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
+```
+## 2.4 Benjamini-Hochberg procedure
 
@@
 
-##2.4
-
 $$ P_{BH}= max_i \left\{ P_{(i)} : P_{(i)} \leq \alpha \frac{i}{m}      \right\} $$
@@
 
-##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false.
-For this method, we use the Holm–Bonferroni method.
-The method is as follows:\\
-
-I have my ordered p-values, $\{ P_1, P_2, ...
-\: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
-Is $P_1 < \alpha/m$?
-so, reject $H_1$ and continue to the next step, otherwise EXIT.
-Is $P_2 < \alpha/(m-1)$?
-so, reject $H_2$ and continue to the next step, otherwise EXIT.
-
-Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$. Using this process, we get the following result:
+##2.5 Holm–Bonferroni method
+
+A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false. For this method, we use the Holm–Bonferroni method. The method is as follows:
+
+I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+
+1.  Is $P_1 < \alpha/m$? Reject $H_1$ and continue to the next step, otherwise EXIT.
+
+2.  Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
+
+3.  Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$.
+
+    Using this process, we get the following result:
 

### 2024-03-09 submission.Rmd
added task describtion

@@
 
+**Task description:**: We first want to study the gene which is the most correlated with the level of cholesterol.
+Identify this gene and plot cholesterol as a function of this gene expression level.
+Fit a linear model.
+Test if there is a significant relation between this gene expression and cholesterol?
+
 We first want to study the gene which is the most correlated with the level of cholesterol.
@@
 ## 2.2 Calculating all p-values
+**Task description:**: We now wish to perform the previous test for all of the 3116 genes.
+For each of the genes, fit a linear model that explains cholesterol as a function of the gene expression and compute the p-value of the test.
 
@@
 ## 2.3 Ordering p-values
+
+**Task description:**Order the p-values and plot the ordered as a function of their rank. On the same plot, display the line y=x/3116. Discuss the result.
 
@@
 ## 2.4 Benjamini-Hochberg procedure
+**Task description:**Identify a set of genes linked to the response (aka discoveries).
+We want to guarantee that the expected proportion of false discoveries (mistakes) is less than 5%.
+Explain how you proceed and how many genes you discover.
 
@@
 ##2.5 Holm–Bonferroni method
+**Task description:** We wish to be more conservative and guarantee that the probability of making a false discovery (or more) is less than 5%. Explain how you proceed and how many genes you discover.
 

### 2024-03-09 submission.Rmd
Added commentary to 1.3

@@
 As observed earlier for a $\alpha=0.05$ we basically have guaranteed that the null-hypothesis is rejected correctly only after 2 samples.
-However, this seems strange and thus one might want to adjust the $\alpha$ value to get a more reasonable sample requirement.
+
+The conclusion is that th Wilcoxon test breaks down with so few samples.
+The Wolcoxon table, which is used to obtain the $\alpha$, does not go below a sample size of 5.
+And in the case that all the values of one distribution is larger, it gives at most an $\alpha$ of 0.1.
+\
+The other case would be to try and fit the data to a distribution. But with so few datapoints, that would be futile. \
+The answer to the question is therefore no, one can not make any conclusions.
 
@@
 ## 2.2 Calculating all p-values
+
 **Task description:**: We now wish to perform the previous test for all of the 3116 genes.
@@
 ```
+
 ## 2.4 Benjamini-Hochberg procedure
+
 **Task description:**Identify a set of genes linked to the response (aka discoveries).
@@
 
-##2.5 Holm–Bonferroni method
-**Task description:** We wish to be more conservative and guarantee that the probability of making a false discovery (or more) is less than 5%. Explain how you proceed and how many genes you discover.
+##2.5 Holm–Bonferroni method **Task description:** We wish to be more conservative and guarantee that the probability of making a false discovery (or more) is less than 5%.
+Explain how you proceed and how many genes you discover.
 

### 2024-03-09 submission.Rmd
feat: final ex1 + 4 version

--- 2024-03-09 submission.Rmd
@@
 
+First we compute the contingency table which we will use to be able to test whether the proportion of diabete males and females are the same.
+
 ```{r}
@@
 
+One method is the Fisher test, which does not rely on sample sizes and can compute p-values exactly. Where we have the following hypotheses:
+
+$$
+H_0: \hspace{5mm} P_{\text{female-diabetic}} =  P_{\text{male-diabetic}} \\
+H_1: \hspace{5mm} P_{\text{female-diabetic}} \neq  P_{\text{male-diabetic}}
+$$
+
 ```{r}
@@
 ```
+
+Indeed we do not observe that the null-hypothesis is rejected.
+However, we do note an odds ratio of around 64.7% thus it is slightly more expected for one group to be assigned the diabete label.
+But again the p-value is not significant enough to draw any conclusions.
+
+Further, one could assume some form of normal approximation of the data and in that case we can apply a 2-sample test for the equality of proportions.
+Which assumes that the difference between the proportions can be drawn from the following normal.
+
+$$
+\hat{p}_A - \hat{p}_B \sim \mathcal{N}\left(0, p(1-p) \left( \dfrac{1}{n_A} + \dfrac{1}{n_B} \right) \right)
+$$
+
+Which we can formulate through the following test.
 
@@
 
+Again we cannot reject the null-hypothesis as the p-value is not significant.
+Another approach which builds on top of this is the Yates continuity correction which subtracts 0.5 from the difference between each observed value to lower the chi-squared value and thus boost the p-value.
+
 ```{r}
@@
 
+As this test implicitly increases the p-value we of course cannot expect to reject the null-hypothesis with this test given the results above.
+Therefore, from the collection of tests above we cannot reject the null-hypothesis and can conclude that the proportion of male and female diabetes are the same.
+
 ## 2. Gene expression data
@@
 
-## 2.1 Finding most significant discovery
-
-We first want to study the gene which is the most correlated with the level of cholesterol.
+### 2.1 We first want to study the gene which is the most correlated with the level of cholesterol.
 Identify this gene and plot `cholesterol` as a function of this gene expression level.
@@
 
-## 2.2 Calculating all p-values
-
-We now do the same test for all 3116 genes and fit a linear model that explains `cholesterol` as a function of the gene expression.
-Doing so gives us a relatively large proportion of genes that correspond to a p-value of less than 0.05
+### 2.2 We now wish to perform the previous test for all of the 3116 genes.
+For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
+AND
+
+### 2.3 Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
 
@@
 
-```
-
-```{r}
-plot(hist(p_list, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
-
-```
-
-## 2.3 Ordering p-values
+p_list=sort(p_list)
+plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
+lines(lineList, col = "red", lwd = 2)
+
+legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
+
+```
 
@@
 In fact, if we were to only have random data, we expect a uniform distribution between 0 and 1 for the p-values.
-Visually, looking at the figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant discoveries.
-
-```{r}
-
-p_list=sort(p_list)
-plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
-lines(lineList, col = "red", lwd = 2)
-
-legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
-
-```
-
-Shuffling all labels, we get an approximate uniform distribution of p-values which corresponds to the p-values following a straight line
+Visually, looking at the Figure above, we can the p-values perform better that the uniform distribution and therefore appear to contain significant findings.
 
@@
 
-plot(hist(p_list_shuffeled, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
-
-```
-## 2.4 Benjamini-Hochberg procedure
+```
 
@@
 
+##2.4
+
 $$ P_{BH}= max_i \left\{ P_{(i)} : P_{(i)} \leq \alpha \frac{i}{m}      \right\} $$
@@
 
+print(index)
+
 plot(p_list, col = "blue", lwd = 2, xlab = "Index", ylab = "p-value")
@@
 
-##2.5 Holm–Bonferroni method
-
-A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false. For this method, we use the Holm–Bonferroni method. The method is as follows:
-
-I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
-
-1.  Is $P_1 < \alpha/m$? Reject $H_1$ and continue to the next step, otherwise EXIT.
-
-2.  Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
-
-3.  Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$.
-
-    Using this process, we get the following result:
+##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false.
+For this method, we use the Holm–Bonferroni method.
+The method is as follows:\\
+
+I have my ordered p-values, $\{ P_1, P_2, ...
+\: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+Is $P_1 < \alpha/m$?
+so, reject $H_1$ and continue to the next step, otherwise EXIT.
+Is $P_2 < \alpha/(m-1)$?
+so, reject $H_2$ and continue to the next step, otherwise EXIT.
+
+Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$. Using this process, we get the following result:
 
@@
 
-Under Louis Bachelier's "Theory of Speculation" assume that the returns can be sampled from a perfectly random distribution, in this case this would be a normal distribution which we can parametirise through a Gaussian model.
+Under Louis Bachelier's "Theory of Speculation" one assumes that the returns can be sampled from a perfectly random distribution, in this case this would be a normal distribution which we can parametirise through a Gaussian model.
 Assume that the return is modeled by some random variable $X_r$ and that some value of that RV is denoted by $x_t$.
@@
 
-Indeed, the it seems that Bachelier's theory does describe the price movements well.
-As the daily return resembles a normal distribution, however, we do observe that that tails seem to follow a higher variance compared to the density around the mean which seems to have a smaller variance.
-Which also explains why our initial Gaussian doesn't effectively capture this density.
+Indeed, the it seems that Bachelier's theory does describe the price movements to some extent.
+However, we do observe that that tails seem to follow a higher variance compared to the density around the mean which seems to have a smaller variance.
+Which also explains why our initial Gaussian doesn't effectively capture this distribution.
 Thus, we can motivate why a Gaussian mixture model might perform better in this scenario.
@@
 
-To combine the multiple Gaussians to model a single valid PDF, we will have to introduce a weight parameter which will be used between the weighted sum of Gaussian components.
+To combine the multiple Gaussians to model a single valid PDF, we will have to introduce a weight parameter which will be used on the weighted sum of Gaussian components.
 First, we denote a single component the following way.
@@
 $$    Y_i = \sum_{k=1}^K \pi_k \mathcal{N}(\theta_k)    $$ As this is a weighted sum and we expect $Y_i$ to be modeled from a valid PDF, we have to constrain the weights such that: $$    \pi_k \geq 0 \hspace{2.5mm} \land \hspace{2.5mm} \sum_{k=1}^K \pi_k = 1    $$ Which is the case as long as we follow that $\pi_k$ comes from a PDF defined over the categorical RVs $Z_i$ which defines the different components such that we have: $$    P(Z_i = k) = \pi_k \hspace{5mm} \text{Where $Z$ is a sequence of RVs} \hspace{5mm} \{1, 2, \dots, K\}    $$ Thus, we observe that we first have to initialize the parameters of the model which we can do by placing the means based on a K-Means initialization and set some explicit variance which will serve as the initialization for the iterative EM-algorithm for parameter optimization over such a GMM.
-First off, we will naively apply a GMM routine to this problem using expectation maximization which we will compare against an constrained optimization approach where we ensure equal means for all distributions as it seems to only be the variance that's needed to change based on our earlier graph.
+First off, we will naively apply a GMM routine to this problem using expectation maximization which we will compare against a constrained optimization approach where we ensure equal means for all components as it seems to only be the variance that's needed to change based on our earlier graph.
 
@@
 We can observe that the final fits are indeed much better than the single Gaussian example.
-However, we do observe that there is some overfitting happening at the tails of the distribution for all of the experiments we have run.
+However, we do observe that there is some over fitting happening at the tails of the distribution for all of the experiments we have run.
 This is caused by the fact that the means are not the same for all the Gaussian components and thus for our next section we will be looking at constrained optimization of this GMM where the means should be the same.
+
+```{r}
+bic <- function(model) {
+  n_free_params <- length(mx2$result$pi)
+  n <- length(model$mixture_EM$x)
+
+  return(-2 * model$log_likelihood + log(n) * n_free_params)
+}
+
+rbind(
+  gmm_2 = bic(mx2),
+  gmm_3 = bic(mx3),
+  gmm_4 = bic(mx4),
+  gmm_5 = bic(mx5),
+  gmm_6 = bic(mx6)
+) %>% as.data.frame() %>% setNames("BIC") %>% rmarkdown::paged_table()
+```
+
+Purely based on the BIC values obtained above we can say that the GMMs with 4 and 5 components are the best due to having the lowest BIC values.
+However, whilst looking at the density plots from above, one can clearly observe that these model overfit the data as it adapts to local variations along the tails of the GMM.
+Thus, a natural progression is to observe how we can improve these GMMs through a constrained optimisation where the means of the components are fixed to be the same.
+Such that we can ignore such local variations.
 
@@
   legend("topright", legend = legend_labels, col = c(rainbow(n_components), "black"), lty = 1, lwd = 2, pch = 20)
-}
-```
-
-```{r}
-constrained_gmm_experiment(y, variances = c(10, 7))
-constrained_gmm_experiment(y, variances = c(10, 7, 5))
-constrained_gmm_experiment(y, variances = c(10, 7, 5, 3))
-constrained_gmm_experiment(y, variances = c(10, 7, 5, 3, 1))
-constrained_gmm_experiment(y, variances = c(12, 10, 7, 5, 3, 1))
-```
+
+  return (out_general=out_general)
+}
+```
+
+```{r}
+mu_mx2 <- constrained_gmm_experiment(y, variances = c(10, 7))
+mu_mx3 <-constrained_gmm_experiment(y, variances = c(10, 7, 5))
+mu_mx4 <-constrained_gmm_experiment(y, variances = c(10, 7, 5, 3))
+mu_mx5 <-constrained_gmm_experiment(y, variances = c(10, 7, 5, 3, 1))
+mu_mx6 <-constrained_gmm_experiment(y, variances = c(12, 10, 7, 5, 3, 1))
+```
+
+```{r}
+mu_mx2
+```
+
+By simply observing the produced densities for the new GMMs it is much better then the standard GMM approach.
+As it quite nicely ignores such local variantions along the tails of the distribution.
+To evaluate the different fits for these Gaussians we perform the BIC tests again.
+
+```{r}
+bic_devicances <- function(model) {
+  n_free_params <- length(out_general$theta$pi)
+  n <- length(y)
+
+  return(model$deviance + log(n) * n_free_params)
+}
+
+rbind(
+  gmm_2 = bic_devicances(mu_mx2),
+  gmm_3 = bic_devicances(mu_mx3),
+  gmm_4 = bic_devicances(mu_mx4),
+  gmm_5 = bic_devicances(mu_mx5),
+  gmm_6 = bic_devicances(mu_mx6)
+) %>% as.data.frame() %>% setNames("BIC (Constrained GMM)") %>% rmarkdown::paged_table()
+
+rbind(
+  gmm_2 = bic(mx2),
+  gmm_3 = bic(mx3),
+  gmm_4 = bic(mx4),
+  gmm_5 = bic(mx5),
+  gmm_6 = bic(mx6)
+) %>% as.data.frame() %>% setNames("BIC (Standard GMM)") %>% rmarkdown::paged_table()
+```
+
+We already see a significant improvement in BIC for the constrained optimization version of the GMM, here the 5 component model is the one with the lowest BIC value and thus the best model in terms of BIC.
+Therefore, this demonstrates clearly the fact that the standard GMM was overfitting to the data and was not able to converge to the optimal solution.
 
@@
 
-Indeed from looking at the above examples the final fits are still not great even when we add additional components to account for the variable variance of the distribution.
+Indeed from looking at the above examples for the standard GMM the final fits are still not great even when we add additional components to account for the variable variance of the distribution.
 But as we observe the distribution it is more clear that it can follow more closely a student t-distribution given its long tails and elongated middle density.
@@
 **Task description:** Propose an algorithm for estimating the parameter of this model and use it to fit this model to the data.
+
+Given that we have the PDF of the student t-distribution above, we want to minimise its negative log-likelihood over the data.
+Which will result in of course maximising the overall likelihood and thus obtain a convergence.
+One can achieve this through an optimiser such as L-BFGS-B which is a gradient based method which computes the gradient $\nabla f(x; \theta)$ where in our case the loss function is defined by the negative log likelihood.
+We perform negative log likelihood due to the fact that this method aims to minimise along the largest negative gradient, and we use log values to avoid gradient vanishing through large float values.
 
@@
 
+As we can observe we obtain a very nice fit for the distribution of the data using the student t-distribution, demonstrating the fact that the data seems to more distributed along a student t-distribution.
+We will in now compare how this model performs wrt the other methods in the following section.
+
 ### 4.5 Model evaluation
@@
 
-By only observing the BIC values it might seem that the single Gaussian is the best model.
-However, this should not be the case, indeed we need to perform additional tests.
-The Kolmogorov-Smirnov test can work for this, and we can expect to see that the single Gaussian is rejected.
-Where we have the following null- and alternative hypothesis
+By only observing the BIC values it is clear that the 5 component GMM constrained on means is the best model.
+Notably we might want to reject the single Gaussian model completely, and we can do this through the Kolmogorov-Smirnov test can work for this.
+Where we have the following null- and alternative hypothesis are defined as follows.
 
@@
 ```{r}
-ks.test(y, "pmixture", out_general_constrained$theta)
-```
-
-```{r}
-ks.test(y, "pmixture", out_general$theta)
-```
-
-```{r}
 ks.test(y, "pmixture", theta0)
@@
 
-```{r}
-# Function to compute CDF of Student's t-distribution
-cdf_student_t <- function(x, nu, mu, sigma) {
-  pt((x - mu) / sigma, df = nu)
-}
-
-# Perform KS test
-ks_test_student_t <- function(y, student_t_params) {
-  # Extract parameters
-  nu <- student_t_params$nu
-  mu <- student_t_params$mu
-  sigma <- student_t_params$sigma
-
-  # Compute CDF values
-  cdf_values <- cdf_student_t(sort(y), nu, mu, sigma)
-
-  # Compute ECDF values
-  ecdf_values <- ecdf(y)(sort(y))
-
-  # Perform KS test
-  ks_result <- ks.test(ecdf_values, cdf_values)
-
-  return(ks_result)
-}
-
-# Perform KS test on the data using the fitted Student's t-distribution parameters
-ks_result <- ks_test_student_t(y, student_t_params)
-ks_result
-
-```
-
-Thus, we can quite confidently conclude that we can reject the single Gaussian as a model (as the null-hypothesis is rejected) and announce the Student t-distribution to be the best distribution given the data.
-Which makes intuitive sense given the fact that the distribution of the data much more resemble that of a Student t-distribution.
+Thus, we can quite confidently conclude that we can reject the single Gaussian as a model (as the null-hypothesis is rejected) and announce the 5 component constrained GMM as the "best" model.
+However, this does not mean that the student t-distribution doesn't model the data well.
 
@@
 **Task description:** Is the expected daily return different than zero?
+
+As we have seen above, a single normal cannot model the data effectively.
+Thus, one will have to use an hypothesis with no normality assumption.
+Which can be done through a Wilcoxon test as we have seen in the previous exercises.
+
+```{r}
+wilcox.test(y, mu=0, alternative="two.sided")
+```
+
+Based on the results of this test, we can reject the null-hypothesis and thus conclude that the expected mean is not equal to 0.
+Which further highlights the fact that the student t-distribution is not the best model for this data given the fact that the mean of a student t-distribution is 0.

### 2024-03-09 submission.Rmd

@@
 
-### 2.1 We first want to study the gene which is the most correlated with the level of cholesterol.
+## 2.1 Finding most significant discovery
+
+**Task description:**: We first want to study the gene which is the most correlated with the level of cholesterol.
+Identify this gene and plot cholesterol as a function of this gene expression level.
+Fit a linear model.
+Test if there is a significant relation between this gene expression and cholesterol?
+
+We first want to study the gene which is the most correlated with the level of cholesterol.
 Identify this gene and plot `cholesterol` as a function of this gene expression level.
@@
 
-### 2.2 We now wish to perform the previous test for all of the 3116 genes.
-For each of the genes, fit a linear model that explains `cholesterol` as a function of the gene expression and compute the p-value of the test.
-AND
-
-### 2.3 Order the p-values and plot the ordered p-values as a function of their rank. On the same plot, display the line $y = x/3116$. Discuss the result.
+## 2.2 Calculating all p-values
+
+**Task description:**: We now wish to perform the previous test for all of the 3116 genes.
+For each of the genes, fit a linear model that explains cholesterol as a function of the gene expression and compute the p-value of the test.
+
+We now do the same test for all 3116 genes and fit a linear model that explains `cholesterol` as a function of the gene expression.
+Doing so gives us a relatively large proportion of genes that correspond to a p-value of less than 0.05
 
@@
 
-p_list=sort(p_list)
-plot(p_list, col = "blue", lwd = 2, xlab = "Rank", ylab = "p-value")
-lines(lineList, col = "red", lwd = 2)
-
-legend("topleft", legend = c("p-values", "Uniform line"), col = c("blue", "red"), lwd = 2)
-
-```
+```
+
+```{r}
+plot(hist(p_list, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
+```
+
+## 2.3 Ordering p-values
+
+**Task description:**Order the p-values and plot the ordered as a function of their rank. On the same plot, display the line y=x/3116. Discuss the result.
 
@@
 
-```
+plot(hist(p_list_shuffeled, breaks=20, col = c("orange", rep("lightblue", 100))), main = "Density of p-values", xlab = "p-values")
+
+```
+
+## 2.4 Benjamini-Hochberg procedure
+
+**Task description:**Identify a set of genes linked to the response (aka discoveries).
+We want to guarantee that the expected proportion of false discoveries (mistakes) is less than 5%.
+Explain how you proceed and how many genes you discover.
 
@@
 
-##2.5 A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false.
-For this method, we use the Holm–Bonferroni method.
-The method is as follows:\\
-
-I have my ordered p-values, $\{ P_1, P_2, ...
-\: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
-Is $P_1 < \alpha/m$?
-so, reject $H_1$ and continue to the next step, otherwise EXIT.
-Is $P_2 < \alpha/(m-1)$?
-so, reject $H_2$ and continue to the next step, otherwise EXIT.
-
-Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$. Using this process, we get the following result:
+##2.5 Holm–Bonferroni method **Task description:** We wish to be more conservative and guarantee that the probability of making a false discovery (or more) is less than 5%.
+Explain how you proceed and how many genes you discover.
+
+A stronger guarantee would be that on impose that there shouldn't be more than a 5 % risk that any discovery is false. For this method, we use the Holm–Bonferroni method. The method is as follows:
+
+I have my ordered p-values, $\{ P_1, P_2, ... \: , P_m \}$, and want the FWER to be no higher than a level $\alpha$.
+
+1.  Is $P_1 < \alpha/m$? Reject $H_1$ and continue to the next step, otherwise EXIT.
+
+2.  Is $P_2 < \alpha/(m-1)$? so, reject $H_2$ and continue to the next step, otherwise EXIT.
+
+3.  Continue this iterative process as long as $P_k < \frac{\alpha}{m+1-k}$.
+
+    Using this process, we get the following result:
 

### 2024-03-09 submission.Rmd
chore: merge

@@
 library(pwr)
-library(mclust)
 ```
@@
 The Wolcoxon table, which is used to obtain the $\alpha$, does not go below a sample size of 5.
-And in the case that all the values of one distribution is larger, it gives at most an $\alpha$ of 0.1.
-\
-The other case would be to try and fit the data to a distribution. But with so few datapoints, that would be futile. \
+And in the case that all the values of one distribution is larger, it gives at most an $\alpha$ of 0.1.\
+The other case would be to try and fit the data to a distribution. But with so few datapoints, that would be futile.\
 The answer to the question is therefore no, one can not make any conclusions.

### 2024-03-10 submission.Rmd
chore: fix in num of params for t-dist
Update submission.Rmd

@@
 
+### 3.1 Data uploading and plotting
+
+**Task description:** Upload the dataset contained in the file data_exo3.csv and plot the data.
+
+The first task asks to read and save the dataset contained in data_exo3.csv.
+
+```{r}
+data = read.csv("data_exo3.csv", sep="t", dec=",", na="")
+```
+
+It's a two-dimensional dataset in two variables denoted as x and y. After this, we want to plot the data on the plane:
+
+```{r}
+plot <- ggplot(data, aes(x = x, y = y)) + 
+  geom_point() + 
+  labs(title = "Scatter plot of the data", x = "X", y = "Y")
+plot
+```
+
+It is evident from the data exploration that observations with lower values of x are situated towards the upper region of the graph, whereas conversely, as x increases, the data points tend to shift lower.
+In the middle range, there is a discernible transitional pattern.
+
+### 3.2 Polynomial Regression Models
+
+**Task description:** Try several polynomial models for the data, select a model and comment the results.
+
+In this section, we embark on the exploration of polynomial regression models.
+Our objective is to discern and select an appropriate polynomial model that best encapsulates the relationship between the variables.
+
+We initiate by fitting polynomial models ranging from degree 1 to 5.
+This choice is motivated by the desire to strike a balance between model simplicity and the ability to flexibly adapt to nuanced patterns within the data.
+
+But first, let's split the data in training and test in order to evaluate the predictive performance of the model.
+
+```{r}
+n <- nrow(data)
+is_train <- sample(c(TRUE, FALSE), n, prob = c(0.8, 0.2), replace = TRUE)
+training <- filter(data,  is_train)
+test     <- filter(data, !is_train)
+```
+
+Starting with the simplest model, a linear regression (degree 1), we aim to assess how well a straight line aligns with the observed data points.
+
+```{r}
+library(ggfortify)
+```
+
+```{r}
+lm1 <- lm(y ~ x, data = training)
+summary(lm1)
+
+predictions  <- 
+  lm(y ~ x, data = training) %>% 
+  predict(test)
+
+pred <- predictions
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+
+plot <-
+  training %>% ggplot() +  aes(x = x, y = y) + 
+  geom_point(size = 2, colour="#993399") +  xlab("x") + ylab("y") +
+  geom_point(data = data.frame(test, testing = predictions),
+             mapping = aes(x = x, y = y), size = 2, shape = 3, colour="red")
+
+plot <- plot + 
+  geom_smooth(method = "lm", data=training, formula = y ~ x, se = FALSE, colour="#339900") 
+plot
+
+autoplot(lm1, which = 1:2) 
+```
+
+The model exhibits a commendable R-squared value of 0.79, signifying a strong fit, and the F-statistic indicates a highly significant overall model (p<0.001).
+The score on the test dataset is slightly lower.
+
+We can see by the diagnostic plots that the model seems to underestimate the data in some intervals and overestimate in others.
+The residuals are not identically distributed around 0, as they follow a trend, and the extreme value are not extreme value of a normal distribution.
+A good idea is considering a polynomial regression of degree 2.
+
+```{r}
+lm2 <- lm(y ~ poly(x, 2), data=training)
+summary(lm2)
+
+predictions  <- 
+  lm(y ~ poly(x, 2), data = training) %>% 
+  predict(test)
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+
+plot <- plot + 
+  geom_smooth(method = "lm", data=training, formula = y ~ poly(x, 2, raw = TRUE), se = FALSE, colour="#177711")
+plot
+
+autoplot(lm2, which = 1:2)
+```
+
+The 2nd degree polynomial model shows a great improvement as it explains 89 percent of the variability of the training data and scores 87 percent on the test data.
+The plot shows how the curve better estimate the data and the diagnostic plot shows how the residuals' trend is closer to the axis.
+
+```{r}
+lm3 <- lm(y ~ poly(x, 3), data=training)
+summary(lm3)
+
+predictions  <- 
+  lm(y ~ poly(x, 3), data = training) %>% 
+  predict(test)
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+
+plot <- plot + 
+  geom_smooth(method = "lm", data=training, formula = y ~ poly(x, 3, raw = TRUE), se = FALSE, colour="#705511")
+plot
+
+autoplot(lm3, which = 1:2)
+```
+
+The 3rd degree polynomial model slightly improves the performance both on the training and the test data.
+On the other hand, there's no confirm on the statistical significance of the 3rd degree component, due to the p-value of 0.16.By the diagnostic plots, we observe that the distribution of residuals is still unbalanced.
+
+```{r}
+lm4 <- lm(y ~ poly(x, 4), data=training)
+summary(lm4)
+
+predictions  <- 
+  lm(y ~ poly(x, 4), data = training) %>% 
+  predict(test)
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+
+plot <- plot + 
+  geom_smooth(method = "lm", data=training, formula = y ~ poly(x, 4, raw = TRUE), se = FALSE, colour="#723366")
+plot
+
+autoplot(lm4, which = 1:2)
+```
+
+The 4th degree polynomial model shows a much better performance, explaining 95 percent of the variability of the data.
+It also has a high predictive power with a 97.7 percent score on the test data.
+The Residual vs Fitted diagnostic plot shows that the trend is closer to zero in this case.
+
+```{r}
+lm5 <- lm(y ~ poly(x, 5), data=training)
+summary(lm5)
+
+predictions  <- 
+  lm(y ~ poly(x, 5), data = training) %>% 
+  predict(test)
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+
+plot <- plot + 
+  geom_smooth(method = "lm", data=training, formula = y ~ poly(x, 5, raw = TRUE), se = FALSE, colour="#720099")
+plot
+
+autoplot(lm5, which = 1:2)
+```
+
+Finally, the fifth grade polynomial model increases slightly the score on the training data from 96 to 97 percent, decreasing the predictive model performance.
+It can be a consequence of overfitting.
+The fifth grade component is not confirmed to be statistically significant due to the p-value of 0.22.
+Overall, it seems less accurate than the previous model.
+
+From the previous observations, we have little reasons to prefer the 4th grade polynomial model. Let's delve into statistical tests in order to quantitatively verify what we observed.
+
+```{r}
+anova(lm1,lm4)
+anova(lm1,lm5)
+anova(lm2,lm4)
+anova(lm3,lm4)
+anova(lm2,lm5)
+anova(lm3,lm5)
+```
+
+As we observed during the single models analysis, the 4th and 5th grade models are highly preferred to the lower grade ones.
+
+```{r}
+anova(lm4,lm5)
+```
+
+By comparing 4th and 5th grade polynomials, the test doesn't let us reject the hypothesis that the 4th grade polynomial is enough to explain the data (instead of using the 5th grade component).
+We can go on with the likelihood ratio test, since out five model are nested with each other.
+
+```{r}
+logLik(lm1)
+logLik(lm2)
+logLik(lm3)
+logLik(lm4)
+logLik(lm5)
+```
+
+The test prefer the 4th and 5th grade polynomials as well, and indicates a little preference for the 4th grade one.
+The difference is not big enough to conclude that the 4th grade model better describes the relations among the variables of the data set.
+We can pursue our investigation by using information criteria
+
+```{r}
+as.matrix(BIC(lm1, lm2, lm3, lm4, lm5))
+as.matrix(AIC(lm1, lm2, lm3, lm4, lm5))
+```
+
+The BIC test leans marginally towards favoring the 4th-grade model, whereas the AIC test yields nearly equivalent values for the 4th and 5th-grade polynomial models.
+
+Considering the inclinations of the AIC and likelihood ratio tests, coupled with the superior prediction score on the test data and the lack of confirmed statistical significance in the 5th-grade component, we adhere to the 4th-grade polynomial regression model as the most suitable polynomial model.
+
+### 3.3 Nonlinear Regression Model and Comparison
+
+**Task description:** Fit a nonlinear model and compare with the polynomial model you have selected.
+
+Given the discernible shape of the data from the plot, a viable approach involves attempting to model the data using the following function:
+y ~ B + A / ( 1 + exp(- gamma * (x -tau)))
+This function incorporates four parameters: A, B, gamma and tau, which are subject to estimation.
+
+```{r}
+library(minpack.lm)
+nlm1 <- nlsLM(y ~  B + A / ( 1 + exp(- gamma * (x -tau))), data = training, start = c(A=0, gamma=0, tau=2, B=2))
+summary(nlm1)
+
+predictions <- predict(nlm1, newdata = test)
+
+R2_test <- cor(predictions, test$y)^2
+print("R2 test: ")
+R2_test
+```
+
+This nonlinear model contains a better description of the data and a quasi-perfect prediction performance with a 99.5 percent score.
+
+```{r}
+new_data <- data.frame(x = seq(min(data$x), max(data$x), length.out = 100))
+
+predictions <- predict(nlm1, newdata = new_data)
+
+library(ggplot2)
+
+plot <-
+  training %>% ggplot() +  aes(x = x, y = y) + 
+  geom_point(size = 2, colour="#993399") +  xlab("x") + ylab("y") +
+  geom_point(data = data.frame(test, testing = pred),
+             mapping = aes(x = x, y = y), size = 2, shape = 3, colour="red")
+
+plot2 <- plot +
+  geom_line(data = new_data, aes(x = x, y = predictions), color = "blue") +
+  labs(title = "Fit del Modello nls", x = "X", y = "Y")
+
+plot2
+
+plot2 +
+  geom_smooth(data = data, 
+    mapping = aes(x = x, y = y),
+    method = "lm", formula = y ~ poly(x, 4, raw = TRUE), se = FALSE, color = "red")
+```
+
+By visual examination, it's evident that the nonlinear model better describes and estimates the data within the range x=3 to x=6.
+It is anticipated that subsequent statistical tests will quantitatively affirm the preference for the nonlinear model when compared to the polynomial counterpart.
+
+```{r}
+logLik(lm4)
+logLik(nlm1)
+```
+
+The likelihood ratio test unequivocally supports the superiority of the nonlinear model, showcasing a substantial advantage over its counterpart.
+
+```{r}
+as.matrix(BIC(lm4, nlm1))
+as.matrix(AIC(lm4, nlm1))
+```
+
+The outcome is affine when evaluating and comparing using information criteria. We have enough evidence to conclude that the preferred model for the description of the dataset is the nonlinear one.
+
+### 3.4 Confidence Intervals and Prediction Intervals
+
+**Task description:** Calculate confidence intervals for E(y_new) and prediction intervals for Y_new on the following values of the covariate: x_new = (1,...,10)
+
+In the context of obtaining prediction intervals and confidence intervals, we employ the delta-method.
+
+```{r}
+# Calculate the derivative
+f_prime <- deriv(y ~  B + A / ( 1 + exp(- gamma * (x -tau))), c("A", "gamma", "tau", "B"), function(A, gamma, tau, B, x){} )
+
+# Generate x_new
+x_new <- seq(1, 10, by=1)
+
+# Obtain the coefficients from the nonlinear model
+beta_hat <- coef(nlm1)
+
+# Calculate the new function values and gradient
+f_new    <- f_prime(beta_hat[1], beta_hat[2], beta_hat[3], beta_hat[4], x_new)
+grad_new <- attr(f_new, "gradient")
+GS <- rowSums((grad_new %*% vcov(nlm1)) * grad_new)
+
+# Set the confidence level
+alpha <- 0.05
+
+# Calculate the confidence intervals and create a data frame
+delta_f <- sqrt(GS) * qt(1-alpha/2,  n - length(beta_hat))
+df_delta <- data.frame(
+  x = x_new, 
+  f = f_new, 
+  lwr_conf = f_new - delta_f, 
+  upr_conf = f_new + delta_f
+  )
+
+# Calculate prediction intervals and add to the data frame
+delta_y    <- sqrt(GS + sigma(nlm1)^2)*qt(1-alpha/2,  n - length(beta_hat))
+df_delta[c("lwr_pred","upr_pred")] <- cbind(f_new - delta_y,f_new + delta_y)
+
+# Display the resulting data frame
+df_delta
+
+# Plot the results
+ggplot(df_delta, aes(x = x, y = f)) +
+  geom_point() +
+  geom_ribbon(aes(ymin = lwr_conf, ymax = upr_conf), fill = "blue", alpha = 0.2, color = NA, linetype = 0) +
+  geom_ribbon(aes(ymin = lwr_pred, ymax = upr_pred), fill = "red", alpha = 0.2, color = NA, linetype = 0) +
+  labs(title = "Confidence Intervals and Prediction Intervals",
+       x = "x_new",
+       y = "y_new") +
+  theme_minimal()
+```
+
+In this visualization, the model's predictions are denoted by black points, accompanied by the respective prediction interval and confidence interval.
+These intervals have been computed for each value of x_new and connected to form two continuous bands encapsulating the forecasted outcomes.
+
 ## 4. S&P500 daily return
@@
 theta = student_t_params
-b_tdist = 3 * log(length(y)) + 2 * neglog_likelihood_student_t(c(theta$nu, theta$mu, theta$sigma), y)
+b_tdist = log(length(y)) + 2 * neglog_likelihood_student_t(c(theta$nu, theta$mu, theta$sigma), y)
 
